Virus specific impacts on honey bee flight performance are mediated by the octopamine pathway

Version 1

Abstract

High annual honey bee colony losses are associated with environmental and biological stressors, including virus infections. In insects, the octopamine pathway orchestrates the “fight-or-flight” response, regulating energy mobilization, temperature, and flight. We determined that sacbrood virus (SBV) infections induce expression of an octopamine receptor and enhance honey flight performance, whereas deformed wing virus (DWV) infections reduce flight performance, but how viruses interface with this pathway remained unknown. To elucidate the relationships between the octopamine response, virus infection, and flight, honey bees were infected with SBV or DWV and exposed to octopamine (OA), epinastine (EP)–an OA receptor antagonist, or both OA and EP; flight and gene expression were assessed. Pharmacologic manipulation revealed that octopamine supplementation rescued flight deficits in DWV-infected bees, but diminished performance in SBV-infected bees, while blocking octopamine receptors altered these effects. Transcriptome analyses indicated that SBV infections, and DWV infection with OA treatment, activated honey bee metabolic pathways, and that SBV infected bees had greater expression of genes involved in OA synthesis, unless treated with OA. These results provide a mechanistic explanation for virus-specific impacts on honey bee flight, which may have consequences on foraging efficiency, colony health and virus transmission.

Introduction

Honey bees (Apis mellifera) are eusocial, generalist pollinators that live in densely populated colonies of ∼30,000 individuals (1, 2). The survival of each colony depends on the coordinated activities of thousands of individual bees that provision nutritional resources, rear brood, and regulate internal hive temperature (3–5). Colony losses in the U.S. averaged 40% from 2008-2025 (6–10), these losses are associated with several abiotic and biotic stressors, including parasites and pathogens (11–14). Honey bee pathogens include bacteria, microsporidia, and viruses (15, 16). Viruses may be transmitted vertically, horizontally, or by vectors, including the Varroa destructor mite (17–21). Varroa mites feed on honey bees and infestations may kill colonies (22). In addition, Varroa mites vector viruses, including deformed wing virus (DWV), which replicates in the mite, and others that are mechanically transmitted such as sacbrood virus (SBV) (23). Most characterized bee infecting viruses have positive single-stranded RNA (+ssRNA) genomes including members of Iflaviridae (e.g., DWV and SBV). Virus infections in honey bees may remain seemingly asymptomatic, cause morphological symptoms, paralysis, and/or result in death (15, 24–28). Honey bees that lack morphological symptoms may still harbor high virus levels that impact lifespan and flight performance (29–32).

Flight is an energetically taxing behavior that is essential for honey bee foraging, defecation, and mating (33–38), and flight muscles are essential for temperature regulation (i.e., fanning to cool and ‘shivering’ for heat) (17, 39–42). In our previous work, we used flight as a proxy for overall bee health and demonstrated that honey bees harboring high levels of DWV without obvious symptoms had impaired flight performance (29). Unexpectedly, we determined that the flight performance of SBV infected honey bees was enhanced and linked to the metabolism-stimulating, stress-induced octopamine pathway (29). Abiotic and biotic stressors may induce enhanced activity of stress pathways including the octopamine response. Invertebrates synthesize octopamine (OA), a ‘fight or flight’ neurohormone analogous to norepinephrine, which facilitates energy production for taxing activities including flight (43–47). The octopamine response may also be triggered by other stressors, including pathogen infections (29, 48, 49).

Whether octopamine signaling is immunosuppressive or immunoenhancing is dependent on multiple factors and its role in honey bee antiviral defense has not been thoroughly investigated (50–52). Honey bee behaviors may be influenced by multiple monoamine neurohormones including OA, serotonin, dopamine, and tyramine (44, 53–58). Octopamine promotes movement, neuromuscular signaling, and metabolism. To synthesize OA, tyrosine is first converted to tyramine by Tyrosine decarboxylase (TDC); tyramine is the precursor neurohormone to OA and is generally associated with reduced flight and movement (45, 59, 60). Tyramine may serve as an antagonistic modulator of behavior to OA or Tyramine β-hydroxylase (TβH) may convert it to OA (Fig 1A). One of the most abundant OA receptors in honey bees is a G-coupled protein receptor, Oβ-2R (61). When OA binds to Oβ-2R, it activates adenylyl cyclase (AC), which increases intracellular concentration of cyclic adenylyl monophosphate (cAMP). Elevated cAMP activates cAMP-dependent Protein kinase A (PKA) which phosphorylates metabolic enzymes and transcription factors that generally enhance metabolic activity (Fig 1B) (40, 46, 61, 62).

Octopamine signaling pathway (i.e., synthesis, activation, inhibition) and experimental design

(A) Octopamine (OA) is synthesized by conversion of tyrosine to tyramine by tyrosine decarboxylase (TDC) and tyramine conversion to OA by tyramine β-hydroxylase (TβH). (B) OA β-2 receptor (Oβ-2R) is one of the most abundant and specific OA-detecting receptors. Upon activation of Oβ-2R, a G-coupled protein receptor, adenylyl cyclase (AC) is activated and increases intracellular cyclic adenylyl monophosphate (cAMP) levels. High cAMP levels induce protein kinase A activity which phosphorylates metabolic enzymes and transcription factors that enhance metabolic activity. (C) An OA receptor antagonist, epinastine (EP), inhibits OA binding (D) To evaluate relationships between virus infection and the OA response, and prevents activation of the Oβ-2R pathway honey bees were inoculated with either deformed wing virus (DWV) or sacbrood virus (SBV) via injection or mock infected. Bees were exposed to either OA, EP, or both OA and EP via co-injection or feeding, all bees were fed sucrose. (E) Impacts of experimental conditions were quantified in individual bees 72 hours post infection. Panels B-D reproduced from Kaku, 2025.

We hypothesized that SBV infections, which correlate with higher expression of an OA receptor (Oβ-2R) (29), activate the OA response which results in enhanced flight performance. Therefore, we predicted that OA treatment would enhance honey bee flight performance regardless of virus infection. Epinastine (EP) is a highly specific OA receptor antagonist that does not appreciably interfere with other neurohormone receptor binding (63, 64). Therefore, we hypothesized that EP treatment would reduce honey bee flight performance and negate SBV-associated enhanced flight performance.

Results

To test these hypotheses, we exposed virus infected honey bees to OA, EP, or both OA and EP and assessed flight performance and gene expression (e.g., tdc, tβh, and oβ-2R) in three independent experiments (i.e., exp- 1, 2, and 3) (Fig 1 and Data S1). Specifically, age-matched adult bees (n = 336) were experimentally infected with DWV (n = 117), SBV (n = 75), or mock infected (buffer injection, n = 144) and treated with OA (10 mM), EP (4 mM), or both OA and EP via injection or feeding, or not treated; all bees were fed 50% sucrose syrup ad libitum.

Treatments of OA and EP were given in 2 µL injections or fed at the same concentration in sucrose syrup; doses were in-line with previous studies (29, 40, 45, 63). Exposure to OA and/or EP via injection was temporally limited, whereas ad libitum feeding provided continuous exposure (Fig 1).

Quantification of experimentally introduced and naturally occurring honey bee virus infections

Honey bees were tested for preexisting infections using pathogen-specific polymerase chain reactions (PCR) and detected pathogens were quantified using quantitative PCR, as described previously (29) (Data S1-S3). We determined that honey bees in exp-1 had preexisting DWV infections and bees in exp-2 and exp-3 had preexisting infections of DWV and SBV (S1 Fig).

Most bees analyzed herein had either experimentally introduced or preexisting DWV infections (n = 325/336) and 40% had experimentally introduced or preexisting SBV infections (n = 134/336). DWV infections ranged between 1×103 and 7.1×1010 copies per 2 µg of RNA (∼7×103 – 8×1011 DWV copies per bee). SBV infections were between 1.2×103 and 1.2×1010 copies per 2 µg of RNA (∼8.3×104 – 8.3×1011 SBV copies per bee). Due to high DWV and SBV prevalence there were no virus-free honey bees. Mock infected bees harbored low virus levels (i.e., an average of 3×105 DWV copies and 2×103 SBV copies per 2 µg RNA) and inoculation resulted in higher virus loads (i.e., an average of 2×108 DWV copies and 2×108 SBV copies per 2 µg RNA), p = 2×10−16 and 5×10−11, respectively). To compare the impact of varying virus levels and treatments, we evaluated several flight metrics (i.e., flight distance, duration, speed, and stop count; Data S1) using linear mixed effect models (LMMs) (complete information for each model and model selections in methods and Data S4).

OA and EP altered flight performance of virus infected bees

The previously described DWV-associated flight impairment relative to mock infected bees (29), was recapitulated in this study (i.e., DWV infected bees flew shorter distances (p = 0.005)). Mock infected bees from this sample cohort, which harbored low virus levels, flew an estimated 45 meters, whereas bees harboring high DWV levels (i.e., 108 DWV RNA copies / 2 µg RNA) flew only ∼9 meters. As expected, DWV-associated flight impairment was alleviated by OA feeding or injection (i.e., average flight distance was similar to mock infected bees fed sucrose, p = 0.61 and 0.08, respectively) (Fig 2A). Honey bees that received EP flew shorter distances (p = 0.002), but the effect of feeding EP was not assessable since bees with experimentally introduced DWV infections that were fed EP died, and there were only 14 bees with low DWV infection levels (Fig 2B, Data S4). Co-injection of OA and EP negatively impacted the distance DWV infected honey bees flew (p = 0.020) (Fig 2C). Counter to our hypothesis, OA-treatment did not enhance flight performance of SBV infected bees (p = 0.69 and 0.60) (Fig 2D, Data S4). SBV infected bees that were treated with EP or co-treated with both EP and OA via injection flew shorter distances than untreated SBV infected bees (p = 0.022 and 0.004, respectively), whereas feeding had no impact (Fig 2E-F).

The flight performance of honey bees infected with DWV and/or SBV is differentially impacted by OA and EP treatment

To evaluate the involvement of the OA pathway on virus infection and/or flight performance, virus- or mock- infected honey bees were exposed to OA, EP, or both OA and EP via co-injection (blue) or feeding (green). Honey bees were experimentally infected with or had preexisting infections of DWV (A-C) or SBV (D-F) in three experiments. Flight distance (log10 m) was evaluated using linear mixed effect models to quantify impacts of DWV and/or SBV infection in the context of OA and/or EP treatment. Each data point represents an individual bee (n = 336); colors indicate treatment (i.e., bees fed only sucrose in pink, bees fed OA and/or EP in green, and bees injected with OA and/or EP in blue). (A) In bees fed only sucrose, DWV infections negatively impacted flight distance, whereas DWV infected bees treated with OA flew further distances. (B) DWV infected bees fed or injected with EP flew shorter distances than DWV infected bees fed only sucrose. (C) DWV infected bees fed or injected with OA and EP flew similar distances to DWV infected bees fed only sucrose. (D) In the control group, SBV infected bees flew greater distances than uninfected bees. SBV infected bees fed OA flew shorter distances than SBV infected bees fed only sucrose. (E) Honey bees with SBV infections fed or injected with EP flew shorter distances than SBV infected bees fed only sucrose. (F) Bees with SBV infections fed or injected with OA and EP resulted in similar flight distances regardless of SBV infection level.

The amount of time a honey bee was actively flying was also assessed as an indicator of energetic capacity (S2 Fig). Honey bees harboring high DWV levels (i.e., >108 DWV copies / 2 µg RNA) flew only 14 minutes (p = 0.022), relative to mock infected bees with low virus infection levels, which flew for 26 minutes. Honey bees with high SBV levels flew for similar durations to mock infected bees (i.e., 28 minutes, p = 0.58), which is consistent with previous results (29). In line with our hypotheses, mock infected bees with low virus infection levels flew longer when they were fed OA (i.e., 49 minutes) (p = 0.0001). DWV-associated flight impairment was mitigated in honey bees that were supplemented with OA, which flew for similar durations to mock infected bees (i.e., 28 minutes, p = 0.76). However, counter to our hypothesis, when SBV infected bees were injected or fed OA, flight duration was significantly reduced (i.e., 18 and 12 minutes, respectively; p = 0.04 and 0.0009).

To quantify the impact of virus and OA and/or EP treatment on honey bee flight speed, we compared average and peak speeds. Average flight speed (i.e., total distance / total flight duration) was similar across all conditions except in bees fed EP, which flew at slower average speeds (p = 3×10−4) (Data S4). We also compared peak speed (i.e., maximum speed reached during flight sustained >10 seconds) (S2 Fig). Similar to previously reported results (29), DWV infected bees had slower peak speeds than mock infected bees (p = 0.016). SBV infected bees flew similar speeds to mock infected bees (p = 0.07), but peak speeds were reduced when SBV infected bees were fed or injected with OA (p = 0.004, p = 0.0006) (Data S4). Mock infected bees flew peak speeds of ∼3.3 km/h, honey bees with high levels of SBV (i.e., 108 SBV copies / 2 µg RNA) flew ∼3.8 km/h, whereas highly SBV infected bees injected or fed OA flew an estimated 2.0 or 2.5 km/h, respectively (p = 0.001 and 0.022).

In addition to flight metrics, post-flight thorax temperatures were measured within a subset of individual bees (n = 130) since flight induces heat and the octopamine response is involved in honey bee temperature regulation (40, 61) (Data S4). The ambient temperature for all flight assays was 22°C, and post-flight thorax temperatures in honey bees maintained in control conditions averaged 28°C. DWV infected bees had similar post-flight temperatures, and highly SBV infected bees were 2°C cooler than control (p = 0.04). Honey bees with DWV infections that were fed OA had thorax temperatures that were 2.5°C higher than bees not fed OA. Whereas the thorax temperatures of bees with DWV infections were 5.6°C lower in bees co-fed OA and EP (p = 0.027 and 1×10−6). Likewise, honey bees with high SBV levels that were injected with OA also had reduced thorax temperatures post-flight (i.e., ∼ 3°C lower) (p = 0.03; Data S4).

Virus specific outcomes of experimentally manipulated octopamine pathway explained by differential expression of key genes

Induction of the OA pathway was previously implicated in the enhanced flight performance of SBV infected honey bees, which had elevated expression of oβ-2R (29). When OA binds to Oβ-2R, it activates AC, increases cAMP concentration, thereby activating cAMP-dependent PKA, which phosphorylates enzymes and transcription factors that enhance metabolic activity (Fig 1B) (40, 46, 61, 62). To determine if DWV, SBV, or different experimental treatments were associated with differences in OA synthesis, we quantified the expression of tdc and tβh relative to mock infected honey bees fed only sucrose (Data S1, Fig 3). We evaluated the relationship between tdc expression and multiple factors including DWV abundance, SBV abundance, experimental treatments (i.e., OA and/or EP, fed/injection) using LMMs which explained 91% of the data variation (Data S4). SBV infection was associated with increased tdc expression (p = 0.021), and SBV infected bees fed OA or co-fed OA and EP also expressed higher tdc levels (p = 0.030 and 2×10−10, respectively). Expression of tdc was lower in SBV infected bees co-injected with EP, and SBV and DWV-coinfected bees that were fed OA or both OA and EP (p-values = 0.033, 0.022, 0.006, 0.004, respectively) (Fig 3A). Expression of tdc was not associated with DWV infection, or DWV infection with OA and/or EP treatment (p-values all >0.2, Data S4). We also examined the effects of virus infection in the context of experimental OA and EP treatments on tβh expression (i.e., in bees with detectable tβh levels n=306/336) (Fig 3B). An LMM that included DWV abundance, SBV abundance, treatment, and tdc expression as explanatory variables explained 85% of the variation in tβh expression (Data S4). Expression of tdc and tβh were positively correlated (with a tdc:tβh ratio of 1:1.6, p = 0.004) (S4 Fig). Expression of tβh was higher in SBV and DWV infected bees (p =0.022 and 1×10−9, respectively), suggesting more tyramine is converted to OA during DWV or SBV infection (Fig 3B). Results from these models indicate that only SBV infections were associated with greater expression of both tdc and tβh relative to mock infected bees with low preexisting virus levels. Whereas SBV infected honey bees that were co-injected with OA expressed similar levels of tdc or tβh as mock infected bees (p = 0.421 and 0.437, respectively). This result suggests that the elevated expression of genes involved in OA synthesis in SBV infected honey bees is reduced by OA supplementation and partially explains the impaired flight performance of these bees relative to untreated SBV infected honey bees (Fig 4).

Expression of tyrosine decarboxylase (tdc) and tyramine β-hydroxylase (tβh) across treatments and SBV infection

The relationships between honey bee tyrosine decarboxylase (tdc) and tyramine β-hydroxylase (tβh) expression in the context of virus infection and treatments were evaluated using a linear mixed effect model that included SBV abundance and OA/EP treatments as fixed effects with an interaction. Individual honey bee data plotted by treatment group with either (A) tdc expression or (B) tβh expression. Gene expression was calculated as log2 fold change using the ΔΔCt method relative to housekeeping rpl8 and compared to mock infected bees. SBV abundance was calculated as log10 copies / 2 µg RNA. The background violin plots represent the 95% confidence interval, and the horizontal line in each violin represents the median. SBV abundance is represented by a green color scale and SBV-negative samples with an “X”. When SBV abundance did not impact tdc or tβh expression the dark green points are evenly distributed above and below the control ‘0’ fold change line, whereas unequal distribution indicates an SBV-specific effect; * p<0.05, **p<0.01, and ***p<0.001. (A) Positive association between SBV abundance and tdc expression (p = 0.021). SBV infected bees fed OA or both OA and EP had greater tdc expression levels than uninfected bees fed only sucrose (p = 0.03 and 2×10−8). (B) Expression of tβh was greater in SBV infected bees than uninfected bees (p =0.022). SBV infected bees fed OA had greater tβh expression levels than uninfected bees fed sucrose (p = 2×10−5). SBV infected bees injected with EP or co-injected with OA and EP had lower tβh expression (p = 2×10−5 and 4×10−9).

Relative tdc and tβh expression in honey bees with high DWV or SBV levels

(A) Tyrosine is converted via tyrosine decarboxylase (TDC) to tyramine. Tyramine is a neurohormone that acts as a behavioral antagonist to octopamine (OA), resulting in reduced movement in invertebrates. Tyramine may act as a signaling neurohormone, or tyramine β-hydroxylase (TβH) may convert it to OA. (B) A linear mixed model was used to assess the relative expression of tdc and tβh in honey bees harboring high levels of either deformed wing virus (DWV) or sacbrood virus (SBV) (i.e., 108 virus copies / 2μg RNA) in the context of OA and/or epinastine (EP) injection. Estimates of relative tdc and tβh expression were generated based on data from 336 honey bees. The single bold dash indicates tdc or tβh expression was similar in virus infected bees and mock infected bees and arrows indicate estimates that were higher or lower than controls.

For OA to stimulate metabolic activity and energy production, OA must activate receptors including Oβ-2R (61). To identify variables that contribute to the range of oβ-2R expression, we compared LMMs including tβh expression, tdc expression, SBV abundance, and OA/EP treatment, which explained 95% of the data variation (S4 Fig). We considered including DWV abundance as a fixed effect in this model, but it did not appreciably contribute to oβ-2R expression (Data S4). These results corroborate our previous findings that SBV, but not DWV, was associated with increased oβ-2R expression (29). The strongest predictors of oβ-2R expression levels were tdc and tβh expression (p = 4.4×10−5 and 3.9 x 10−5) and SBV was positively associated with tdc and tβh expression (p = 0.0006) (Data S4).

Transcriptome level impacts of virus infection, flight, and OA treatment

We hypothesized that flight-impacting virus infections would result in differential expression of genes involved in stress, metabolic, and immune pathways. To test this, the transcriptomes of SBV, DWV, DWV and OA, or mock infected honey bees were compared. Honey bees were obtained from a previous experiment and included individuals that did or did not fly (29). RNA was prepared for high throughput sequencing on an Illumina NovaSeq X Plus 25B (n = 36 total individual bee samples, 3-5 per treatment group; Data S6). An average of 25 million reads was obtained for each sequencing library and mapped to the A. mellifera assembly HAv3.1 (65, 66). To evaluate similarities and differences between the transcriptional responses of honey bees with flight-impacting virus infections, we compared DWV, SBV, and DWV and OA injected bees relative to mock infected bees immediately post-flight since these bees had little to no detectable preexisting infections, whereas the no-flight mock infected bees had preexisting DWV infections (Data S5-S6).

Hundreds of differentially expressed genes (DEGs) were identified in SBV, DWV, and OA treated DWV infected bees (Fig 5). DWV infected bees with and without OA treatment shared 95 DEGs, which was more than they shared with SBV infected bees, indicating that there are virus specific transcriptional responses. There were 24 genes shared among SBV, DWV, and DWV and OA injected bees including elevated expression of those involved in the Ras-Raf-MEK-ERK pathway (including mesr3 and rap2l) which is associated with metabolism, muscle development, and the cAMP/PKA pathway (67). Genes with lower expression were involved in muscle function (i.e., mlc2, mlc1, cher, sar), energy production (i.e., cnn, ant, pka-c3), and an inhibitor of proteolytic activity (i.e., cyp4g11).

Transcriptome level comparison of honey bees that were SBV infected, DWV infected, or DWV infected and OA-treated

Gene expression was evaluated in honey bees that were mock infected, SBV infected, DWV infected, or DWV infected bees that were co-injected with OA (DWV+OA) after flight. Differentially expressed genes (DEGs) for each treatment group were identified relative to mock infected bees and compared using a Venn diagram. Full list of DEGs, fold change, and Benjamini-Hochberg corrected significance values are included in SI Appendix Table S6.

There were many genes unique to SBV and DWV infections (Fig 5). Honey bees with SBV infections exhibited lower expression of several genes associated with the citric acid cycle (acsl6, cg5065, acyl-coA delta(11), mcad, nd-13a, nd-13b, nc73ef, and mdh2) and general muscle function and tissue repair genes (tmp-1, tmod, plastin-2, and fgf17). SBV infected bees had elevated expression of rgs11–which increases G-protein α-subunit active state to inactive to expedite the G-coupled protein response (68), titin–which contracts stretched/striated muscle tissues (69), and several immune genes such as argonaute-2, relish, and domeless (Data S6-S7). DWV infected bees exhibited greater expression of several genes associated with immune function including protein mahjong, spastin, and a small heat shock protein, protein-lethal(2) essential for life-like (Data S6). The DEGs unique to DWV infected bees post-flight that exhibited lower expression included genes involved with muscle function and structure (obscurin, a titin homolog, ryr, actin, mhc, stacl, tpm1, and zasp66) and ATP generation (citrate synthase, adenylate kinase 1, ADP/ATP translocase).

Since OA supplementation mitigated DWV-associated flight impairment, we identified differences associated with OA by comparing DEGs to untreated DWV infected honey bees post-flight. As hypothesized, DWV infected bees that were co-injected with OA treatment had greater expression of OA pathway genes (oβ-2R, adenylyl cyclase (ac), and CREB binding proteins) and genes involved in glucose metabolism (gale, gdh, adpgk, putative glucosylceramidase 4, gba1lp, g6pd, sugarless, pdp, ND-24, NAD kinase, tret1, 1,5-anhydro-D-fructose reductase, and igfals) (Data S8-S9). These bees also expressed higher levels of genes induced by high cyclic guanine monophosphate (cGMP) levels (i.e., burgundy, cGMP-dependent protein kinase-like, and cGMP-dependent protein kinase 1) and genes in the RAS/MAPK-pathway (rho1, rac1, and rab5). Untreated, DWV infected bees expressed higher levels of several immune genes (abaecin, domeless, echinoid, and defensin-1) and heat shock protein encoding genes (hsp83, hsp90, hspβ-1) relative to DWV infected bees treated with OA (Data S7-S9).

After flight, both SBV infected and OA treated DWV infected bees expressed lower levels of several mitochondrial associated genes (vdac, blw, atpsynβ, cox10, klf6, and mcmp1) and genes involved in energy generation (pyruvate kinase, gapdh2, and 2 ND-49, and ND-51) than mock infected bees (Fig 5) This result suggested that flight was more taxing for virus infected bees. In addition, transcriptome level results indicated that OA treatment during DWV inoculation resulted in enhanced energy production, since these bees expressed higher levels of genes involved in energy production post-flight relative to DWV infected bees that did not fly (Data S10).

Since OA treatment resulted in recovered flight performance in DWV infected bees, we compared the transcriptomes of honey bees with similar DWV-levels that were either OA-treated or untreated. The 50 genes with the greatest difference in mean expression were identified in bees that did or did not fly (i.e., 100 total possible genes) (Data S10). There were 30 shared genes associated with OA-treatment regardless of flight status. This list included defensin-1, apidermin-2, melittin, odorant binding protein 3, chymotrypsin-1, two α-glucosidases, peritrophin, 2 chitinase proteins, fibrillin-2, muscle protein 20, and 7 uncharacterized genes. The OA-associated list also included vegetative cell wallprotein-like (gpl-1), which exhibited higher expression in bees that flew further distances (mock, SBV, DWV and OA), than in DWV infected bees. Previous studies determined that gpl-1 expression was greater in Nosema ceranae infected honey bees (70), and that octopamine signaling was also impacted (48, 49). While the bees in our study were not Nosema infected (S1 Fig), we used qPCR to confirm that bees expressed higher gpl-1 levels when injected with OA or EP (LMM, p = 2×10−7 and 4×10−8, respectively); data from additional experiments (i.e., exp-1, n = 61 and exp-2, n = 69; Data S1). We also examined expression of a subset of OA-associated genes including OA receptors (i.e., oα-1R, oβ-1R, oβ-2R, oβ-3R), enzymes involved in OA synthesis (tdc and tβh), and genes associated with Oβ-2R activation (adenylyl cyclase (ac), protein kinase A (pka), and CREB-associated protein encoding genes) (Data S9 and S5 Fig). Expression of tdc, tβh, oβ-1R, crebA, and creb-binding isoform X1 was generally higher in SBV infected honey bees, and greater tdc and tβh expression was confirmed via qPCR (exp 1-3, n = 336) (Fig 3, Data S1 and S4). Honey bees that were simultaneously DWV infected and treated with OA expressed higher levels of oβ-2R, oα-1R, and creb binding isoform X2 than untreated DWV infected bees (S5 Fig); greater oβ-2R expression was also confirmed by qPCR (LMM, n = 336; p = 9×10−4; Data S1). Expression of oa-1R and tyrR was assessed via qPCR in only one experiment since expression was generally low regardless of treatment (exp-3, n = 69; Data S1). SBV infected bees had greater expression of oa-1R (p = 0.017), and SBV infected bees co-injected with EP had significantly lower expression of both oa-1R and tyrR (p = 0.0006 and 0.04, respectively).

Discussion

The interactions between virus infections, host immune and stress responses, and organismal health are complex. Insects rely on OA as a neurohormone and neuromodulator to regulate physiological processes and behaviors (43, 45, 71–73), and our previous studies indicated that viruses differentially impacted honey bee octopamine signaling and flight performance. Therefore, we used honey bees as a model system to further investigate these interactions at organismal and transcriptional levels.

As previously described and shown herein, inapparent (or asymptomatic) DWV infections negatively impact honey bee flight performance (i.e., flight distance, duration, and peak flight speed). Evaluation of transcriptional responses to DWV infection suggest that DWV-associated flight impairment may be partially explained by lower expression of genes involved in energy production, which were elevated in OA treated honey bees (Fig 5, S5 Fig, Data S7-S8). These results are supported by previous studies that showed DWV infected honey bees expressed lower levels of G-coupled protein receptor encoding genes, as well as other genes involved in metabolism (74–76). Previously, we demonstrated that the flight performance of SBV infected honey bees was enhanced and that these bees expressed more oβ-2R (29). Herein, we showed that SBV-associated flight enhancement was reduced when OA signaling was inhibited by EP, an OA receptor antagonist (Fig 2 and 4). We expected that SBV infected bees would have compounding, enhanced flight performance when treated with OA, but unexpectedly, OA supplementation negatively impacted flight performance (Fig 2). To further investigate this physiological response, we examined the expression key genes in the octopamine pathway and determined that SBV infected honey bees expressed greater levels of tdc, tβh, and an OA receptor, oβ-2R. Whereas, expression of tdc and tβh was lower in SBV infected bees that were treated with OA (Data S1 and S4, Fig 3-4). Reduced expression of these genes would likely limit OA synthesis, OA signaling, reduce metabolic activity, and result in less available energy for flight (43). Therefore, the reduced flight performance of SBV infected bees treated with OA may be explained by a feedback loop that regulates intracellular OA concentrations.

Depending on OA concentration and specific pathogen challenge, OA may enhance or inhibit immune function in many invertebrates, though mechanistic studies have been primarily restricted to two model organisms (51, 52, 77–79). In Drosophila, OA treatment impacts intracellular OA concentration, likely in a feedback mechanism that regulates neurohormonal signaling (56, 80, 81). In Caenorhabditis elegans, OA pathways are hypothesized to be constitutively active, but downregulated upon pathogen infection to enable an effective immune response while limiting immune activity in the absence of pathogen challenge (77).

Transcriptomic analysis of honey bees infected with a model virus (Sindbis) that were co-injected with double-stranded RNA, a virus-associated molecular pattern, had greater expression of an OA receptor (oβ-2R) (82). Results presented herein indicate that OA treatment of DWV infected honey bees alters expression of several genes associated with immune function including apidermin, defensin, and melittin (83–86). Additionally, DWV infected bees with OA treatment had reduced expression of an odorant binding protein that was previously associated with SBV infection, indicating that it may be involved in honey bee antiviral defense (87). Together, these results highlight virus-specific transcriptional responses in honey bees that may impact virus dynamics and flight performance during commonly occurring coinfections (29, 88).

Multiple OA receptors have been identified in honey bee flight muscles and neural tissues including OαR1, Oβ-2R, Oβ-1R, and Oβ-3/4R (40). OA receptors have also been detected on invertebrate immune cells (51). In addition to their involvement in flight, OA receptors are also involved in honey bee hygienic, communicative dance, and thermoregulatory behaviors (40, 53, 57, 61, 89). Typically around 12 days post-emergence honey bees change from nurse bees, which care for developing brood, to pre-forager bees and undergo physiological changes required for flight (90, 91). As honey bees age and become foragers, OA levels in the brain increase (44, 55, 92). Honey bees infected with the microsporidian parasite, Nosema, have greater OA levels than uninfected bees (48, 49) and are prone to precocious foraging (93), though Nosema levels did not appreciably impact honey bee flight capability (30). These findings are in-line with SBV-induced flight enhancement that correlated with higher oβ-2R expression levels (29). Previous laboratory-based studies demonstrated that precocious foraging may be induced by OA supplementation (44).

Honey bees are exposed to environmental OA sources including citrus blossom nectar, which may contain tyramine and OA at concentrations up to 256 µM (94–97). In addition, acaracides that are OA receptor agonists including amitraz and chlordimeform are commonly used in agriculture (98). Amitraz is commonly used to mitigate Varroa mite infestations in honey bee colonies (99). Varroa mites with amitraz-resistant Oβ-2R mutations have been reported (100, 101). Amitraz has generally been considered safe for honey bees since honey bee OA receptors are divergent from Varroa mite OA receptors (i.e., 48% identical at the amino acid level) (99, 102). However, honey bees exposed to amitraz had higher mortality when co-stressed via infection with a model virus (Flock House virus) (103). These results in conjunction with our findings, linking virus infection with the OA stress response pathway, may indicate previously unrecognized impacts at the organismal level. Varroa mites vector many honey bee viruses and infested colonies often have coinfections and greater average virus levels (104).

Therefore, application of miticide to reduce Varroa and control vectored viruses, or the accumulation miticide metabolites in wax or other hive matrices, may inadvertently harm co-stressed, virus infected honey bees (103, 105, 106). Collectively, the data presented herein demonstrate that the OA pathway is critical for mediating virus-specific effects on honey bee flight performance, which may in turn influence viral transmission and colony health.

Materials & Methods

Honey bees

Honey bees (Apis mellifera, primarily carnica), were maintained with standard apicultural practices at Montana State University’s Horticulture Farm and bees were collected 24 hours prior to eclosion for each experiment. Each experiment was performed with honey bees from different colonies to obtain robust results from genetically diverse bees (Data S1). Bees were cohoused in modified deli containers (n = 10 per house), maintained at 32°C in an incubator and fed 50% sucrose syrup and water ad libitum (107). All bees were fed only sucrose syrup, unless octopamine (OA) and/or epinastine (EP) oral treatment was otherwise specified. OA was administered at a concentration of 10 mM and EP at 4 mM in fresh sucrose syrup daily, and honey bees were given unrestricted access to the supplemented sucrose syrup for the duration of the experiment.

Virus stock preparation

Deformed wing virus (DWV) and sacbrood virus (SBV) inoculums were propagated in white-eyed pupae as previously described (29). In brief, pupae were injected with 2µL of Tris-HCl containing either 4.0×106 DWV RNA copies or 3.5 x 104 SBV RNA, between the 2nd-3rd abdominal tergites using a Harbo syringe and a disposable borosilicate needle made from modified capillary tubes. Pupae were maintained in individual wells in a 24-well plate with a piece of Whatman filter paper at the bottom of each well to prevent pupal tissue damage. In each plate, four wells were filled with sterile water for humidity, at 32°C for 7 days. Pupae were then homogenized in a 2 mL microfuge tube with a sterile, metal, 3 mm bead in 1 mL 1X PBS (pH 7.4) using the TissueLyser II (Qiagen) at 30 Hz for 2 min. Homogenates were centrifuged at 14,000 x g for 15 min at 4°C and supernatant was transferred to a fresh tube. The virus-containing lysates were filtered through a 0.45 µm filter, then a 0.22 µm filter to remove large particles and non-viral microbial contaminants, respectively. RNA was isolated from the virus-containing filtrate (20 µL) using TRIzol (Thermo) according to manufacturer’s instructions. RNA was quantified using a NanoDrop 2000c Spectrophotometer, cDNA was synthesized with reverse transcriptase, and virus RNA copies (including both genomes and transcripts) were quantified from cDNA using qPCR. The inoculum was tested for copurifying/contaminating viruses via PCR (Data S2, S2 Fig). RNA copy concentration in the purified virus stock was quantified relative to a standard curve using qPCR (Data S3). DWV genome consensus sequences were assembled from short-read sequencing data obtained from sequencing RNA isolated from the inoculum (DWV-lab 2024; GenBank PV821422) and SBV sequencing data (sacbrood virus isolate MT 2023; GenBank PV788228.1). Virus inoculum aliquots were stored at −80°C until each experiment.

Experimentally introduced honey bee virus infections

Honey bees were infected with DWV using previously described methods (107). In brief, age-matched bees were cold anesthetized at 4°C for 10 min and infected via intra-thoracic injection using a Harbo syringe with 3.5 x 104 DWV RNA copies, 3.5 x 104 SBV RNA copies, or coinfected with 3.5 x 104 DWV RNA copies and 9.1×103 SBV RNA copies. If octopamine (OA) and/or epinastine (EP) treatment was administered via injection, the injection inoculum also contained 10 mM OA and/or 4 mM EP so that each bee only experienced one injection to reduce injection-associated stress. Each inoculum dose was suspended in 2 µL10 mM Tris HCl buffer, pH 7.5. Buffer injected/ mock infected bees were injected with 2 μL buffer. The DWV doses utilized in this study were similar to previous studies and resulted in virus levels commonly found in naturally infected bees (8, 30, 108–111). As previously described (29), bees had #000 steel washers affixed to their thoraces using rubber cement immediately after injections; washers had an outside diameter of 0.19 cm and weighed 4.3 mg, less than half of an average pollen load (112).

Flight mill

The flight mills used in this study were based on flight mills used in previous honey bee studies (29, 39, 113, 114) and included a magnet to tether bees from the affixed steel washer on each thorax, which enabled their removal from flight mills with minimal damage. Flight assays were performed as previously described (29). In brief, bees 72 hpi (i.e., 4-days post eclosion) were immobilized via incubation at 4°C for ∼12 minutes and tethered to flight mills; counterweights were adjusted for each bee. Temperature and humidity were maintained the same for each experiment (23°C and 31±4% relative humidity). Flight was instigated by tapping the flight arm downward and was re-initiated up to three times before bees were considered ‘exhausted’ and the data collected was representative of ‘total flight’ capability (Data S1). Bees were stored at - 80°C immediately after flight. Flight distance, duration, and speed calculations were based on redlight detector counts which were recorded on an attached Raspberry Pi as previously described (29). Active flight duration was calculated as the sum time moving >5.5 cm (one encoder count) (Data S1). Flight stop count was defined as any stop that exceeded 1.5 seconds. Honey bees may stop to seemingly engage in grooming behaviors, stops were counted if the bee re-initiated flight (29). DWV infected, EP injected, SBV infected bees fed OA or EP or injected with OA, or SBV infected bees fed both OA and EP had fewer stops than mock infected bees (Data S4).

Honey bee RNA isolation

RNA was isolated from the abdomen of each individual honey bee since abdomens were representative of the entire bee (29). Each abdomen was individually homogenized in sterile water (300 μL) with one sterile steel bead (4.5 mm) using a Qiagen TissueLyser for 2 min at 30 Hz. Lysates were centrifuged at 4°C for 5 minutes at 7,500xg. RNA was isolated from the supernatants using equal volumes of TRIzol reagent according to manufacturer’s instructions. The concentration and quality of RNA samples were assessed on a ThermoFisher Nanodrop 2000 spectrophotometer. RNA was stored at −80°C.

Reverse transcription / cDNA synthesis

Reverse transcription reactions were performed by incubating 2 μg RNA, 200 U M-MLV reverse-transcriptase and 500 ng random hexamer primers in 25 μL reactions for 1h at 37°C, according to the manufacturer’s instructions. cDNA was diluted with sterile water (1:2) and 2 μL were used for polymerase chain reactions (PCR).

Polymerase Chain Reaction

Since bees were obtained from colonies that are subject to natural infections (107, 108), PCR was performed to test for preexisting virus infections. Specifically, 2 μL cDNA template was combined with 10 pmol of each forward and reverse primers (Data S2) and amplified with ChoiceTaq polymerase according to the manufacturer’s instructions using the following conditions: 95°C for 5min; 95°C for 30s, 57°C for 30s, 72°C for 30s, 35 cycles, followed by final elongation at 72°C for 4min. PCR products were assessed by gel electrophoresis (1.5% agarose with SYBR safe) and visualized using a Syngene U:Genius 3 imaging system (S1 Fig). PCR products were previously verified by sequencing (108). Analysis of pooled cDNA samples from mock infected honey bees from each of these experiments revealed that many bees had preexisting DWV and SBV infections, therefore qPCR was used to assess their abundance in all individual samples (Data S1). The qPCR primers used in this study to target DWV, bind to both DWV-A (GenBank AY292384) and DWV-B (GenBank MN565036), in addition to numerous other DWV sequences on NCBI including the consensus sequences of the DWV inoculum as part of this study (Datal S3).

Quantitative PCR

Quantitative PCR (qPCR) was used to assess virus abundance and relative abundance of honey bee transcripts. All qPCR reactions were performed in triplicate with 2 μL of 1:2 diluted cDNA template. Each reaction contained 1× ChoiceTaq Mastermix, 0.4 μM each forward and reverse primer, 1× SYBR Green (Life Technologies), and 3 mM MgCl2 for a total volume of 20 μL per reaction. A CFX Connect Real Time instrument (BioRad) was used for the following: pre-incubation 95°C for 1 minute followed by 40 cycles of 95°C for 10s, 58°C for 20s, and 72°C for 15s, with a final melt curve analysis at 65°C for 5s to 95°C. To quantify virus copies in each sample, plasmid standards (i.e., virus-specific PCR amplicons cloned into the pGEM-T (Promega) vector and sequence verified) for each virus were used as templates, ranging from 103-109 copies per reaction for each standard curve as previously described (Data S3) (29). The honey bee gene rpl8 was used as a housekeeping gene for each sample for comparison and for relative fold change calculations (Data S1-S2). Sterile water containing no DNA template was used for negative controls. qPCR specificity was verified by melt point analysis, gel electrophoresis, and sequencing (108). The starting quantity (SQ) of cDNA template was calculated for each honey bee sample as previously described (29). For each honey bee sample, the starting quantity (SQ) of cDNA template (representing 80 ng total RNA) was calculated based on the standard curve and subtracting the average SQ of the no template control reaction (<600 copies). Samples with virus RNA copies below the limit of detection for qPCR (<1000 copies / 2 µg RNA) were listed as 0 (Data S1). Virus abundance was reported as virus RNA copies, including genomes and transcripts, per 2 µg total RNA (i.e., per RT reaction) and ranged up to 7×1010 virus RNA copies / 2µg RNA. Whole bee data was estimated based on the previous estimate of 138 µg RNA per whole bee (29). Comparison of virus abundance levels using estimated copy numbers enabled inclusion of values of 0. Using the ΔΔCt method to calculate virus levels in bees with no detectable virus would inaccurately result in ‘missing’ data and consequently skew model interpretations. Relative host gene expression was calculated using the average ΔΔCt method where ΔCt was determined by subtracting rpl8 Ct from the gene of interest Ct. Average ΔCt was calculated by the mean ΔCt of the buffer-injected bees in each experiment, which harbored the lowest virus levels. The ΔΔCt was calculated by subtracting the average control ΔCt and fold change was determined by the equation 2−ΔΔCt (Data S3).

RNA sequencing and analysis

Honey bee samples with low/no preexisting infections (n = 3-5) were selected per treatment from previously published data (29) (n=36 total) (Data S5). RNA from each sample was DNase treated using Qiagen RNeasy columns. RNA samples were sent to the Roy J. Carver Biotechnology Center at University of Illinois; quality was assessed with an Agilent 3500 fragment analyzer. Libraries were prepared via Watchmaker polyA RNA-Seq prep kit according to manufacturer instructions. Libraries were sequenced on an Illumina NovaSeq X Plus 25B (paired 150 nt reads), approximately 3.2 billion total paired end reads were obtained.

Sequencing quality of reads representing individual samples using FastQC, and filtering of the comprehensive dataset was done using MultiQC versions 0.12.1 (115). Illumina adaptors were trimmed and low quality reads were omitted when Q<30 with BBduk from BBTools version 39.26 (116). Reads were normalized with shrinkage estimations for dispersion using DESeq2 l (correction with q-value<0.05) (117). Sequencing libraries were aligned to the Apis mellifera genome, Amel_HAv3.1 (i.e, GCA_003254395.2_Amel_HAv3.1_genomic.fna, downloaded from NCBI 2019–02-07) (65, 66). Sequencing data from individual samples were uploaded to the NCBI short read archive (SRA) (BioProject ID: PRJNA1414634). Genes were considered differentially expressed if log2 fold change of transcript abundance had adjusted p-values <0.05. Venn diagrams were produced using the eulerr package in R (118).

For virus quantification, trimmed paired end read files were aligned to a nonredundant version of the Holobee database (i.e., https://doi.org/10.15482/USDA.ADC/1255217) (119) which considered 32 potential virus alignments, using HISAT2 version 2.2.1 (120). Default parameters were used during alignment to allow multimapping of reads. Samples with unexpected infections were omitted from analyses (n = 6/36; Data S5). Alignment files were converted to compressed alignment files and sorted using samtools version 1.21 and unmapped reads were discarded (121). The samtools built-in idxstats command was used to quantify virus abundance in sorted alignment files and generate read counts. Read counts of each trimmed FastQ file were quantified using the Seqkit stats command (122). Computational efforts were performed on the Tempest High Performance Computing System, operated and supported by University Information Technology Research Cyberinfrastructure (RRID:SCR_026229) at Montana State University.

Statistics

Statistical analyses were performed in R 4.3.3. Log-transformations and non-parametric tests were performed as necessary. Specifically, all relative gene expression data was log2 transformed and all virus abundance data were log10(1+) transformed, which enabled inclusion of 0s. All mixed models were assessed using the lmer4 package (123). Assumptions for linear models were verified by diagnostic plots and histograms using base R functions and the performance package (124, 125). All models included experiment ID as a random effect and where relevant, virus injections and flight mill IDs were also included as random effects (Data S4). Models were compared using corrected Aikike information criterion (AICc) and Bayesian information criterion (BIC). Models were assessed via maximum likelihood (ML) when random effects varied and via restricted maximum likelihood (REML) when fixed effects varied. We reported only results from the best model for each analysis, but other considered models were included in supplementary material (Data S4). For each linear mixed model with reported estimates, p-values were estimated from t-statistics via Satterthwaite’s method (126).

Figures

All figures were generated through GraphPad Prism software for windows using version 10.6.0, in RStudio using version R 4.3.3, in BioRender, and Adobe Illustrator.

eLife Assessment

This important study investigates how distinct honey bee viruses differentially alter flight performance through interactions with octopamine signaling pathways. The combination of behavioral flight assays, pharmacological perturbation, and transcriptomic analyses provides solid evidence that virus-specific effects on flight are associated with octopamine signaling. However, some of the stronger mechanistic conclusions regarding direct regulation of octopamine signaling remain incomplete without more specific validation of receptor-level effects and direct quantification of octopamine levels or signaling activity.

Reviewer #1 (Public review):

Summary:

Kaku and Flenniken investigate the mechanistic pathways through which specific viral infections alter the flight capabilities of honey bees. Building on their previous discovery that DWV impairs flight while SBV unexpectedly enhances it, the authors hypothesized that these behavioral shifts are driven by interactions with the insect's octopamine (OA) signaling pathway, which is responsible for the "fight-or-flight" neurohormonal stress response and energy mobilization. To test this, the authors experimentally infected adult honey bees with DWV or SBV and pharmacologically manipulated the OA pathway using either octopamine supplementation or epinastine (EP), an OA-receptor antagonist. They then evaluated the bees' flight performance (distance, duration, and speed) on custom flight mills and profiled their gene expression using qPCR and RNA sequencing.

Strengths:

A major strength of this study is the high prevalence of preexisting background DWV and SBV infections in the honey bee cohorts, which meant there were no completely "virus-free" control groups. However, the authors successfully mitigated this limitation by rigorously quantifying viral RNA copies for every individual bee via qPCR and utilizing these viral abundances as continuous variables in powerful linear mixed-effect models.

Weaknesses:

The primary weakness lies in the methodology used for targeted pharmacological manipulations, as well as the lack of OA quantification across different treatments. Thus, their claims are not sufficiently supported by the current data.

(1) The authors utilize Epinastine to block octopamine signaling, describing it as a highly specific OA receptor antagonist. However, pharmacological inhibitors often lack absolute specificity. Epinastine might bind to other octopamine receptor subtypes present in honey bee neural and flight muscle tissues, or it could potentially cross-react with tyramine and dopamine receptors. Without further genetic validation (e.g., RNA interference targeting specific receptors), it is difficult to definitively conclude that the altered flight performance is solely due to the blockade of the specific Oβ−2R pathway.

(2) As a natural neurotransmitter, insects have evolved highly efficient "cleanup" mechanisms. OA is rapidly cleared from the synaptic cleft via reuptake transporters and quickly inactivated by enzymes such as N-acetyltransferase (NAT) or Monoamine Oxidase (MAO). Consequently, an injection of OA produces only a transient "pulse" of activity. It is often a poor "tool" for inducing prolonged physiological effects compared to synthetic formamidines like Amitraz.

(3) The study relies heavily on transcriptomics and quantitative PCR to measure the mRNA expression of key synthesizing enzymes, namely tyrosine decarboxylase (tdc) and tyramine β-hydroxylase (tβh), to infer the activation or suppression of the octopamine pathway. However, changes in enzyme synthesis at the RNA level are often insufficient to accurately reflect the true physiological levels of biogenic amines. To robustly prove the authors' hypothesis of a "feedback loop that regulates intracellular OA concentrations", direct quantification of actual octopamine and tyramine titers in the bees (e.g., using high-performance liquid chromatography or mass spectrometry) is necessary.

Reviewer #2 (Public review):

Summary:

This highly original and well-designed study provides insight into how honeybee picorna-like viruses, Deformed wing virus ( DWV) and Sacbrood virus (SBV), affect flight performance, and reveals the role of the octopamine (OA) pathway in virus-honeybee interactions. The authors used a flight mill to quantify the flight performance of bees with different levels of DWV and SBV. Bees were treated with OA and/or epinastine (EP) - an OA receptor antagonist; the study also quantified virus loads and expression of two key genes involved in OA biosynthesis.

The results showed that reduced flight performance associated with high DWV levels could be alleviated by OA administration. In contrast, increased levels of SBV had the opposite effect, leading to enhanced flight performance. This suggests distinct physiological responses to DWV and SBV infections. Administration of EP had led to a reduction of flight performance in SBV-infected bees, indicating the involvement of the OA pathway.

The authors also quantified levels of mRNAs of enzymes involved in OA synthesis, tyrosine decarboxylase (TDC) and tyramine beta-hydroxylase (TbH), and concluded that DWV induced expression of TbH, while SBV upregulated expression of TDC. Furthermore, the study identified upregulated and downregulated genes in response to SBV, DWV and DWV in combination with OA.

Strengths:

The study reported opposing effects of infections of related viruses, SBV and DWV, on honeybee flight performance, and identified the central role of the octopamine (OA) signaling pathway in the effect of viruses on honeybee flights.

These findings were achieved by using a combination of approaches, including experimental measurement of flight distance, virus infections, and introduction of OA and EP. Experimental work with honeybees is technically challenging and requires specialized expertise, which makes the results produced in this study more valuable.

DWV and SBV are among the most important honeybee pathogens affecting honeybee health and threatening the pollination service. Therefore, an understanding of the mechanisms underlying DWV and SBV pathogenesis has the potential to develop novel approaches to mitigate the negative impact of these viruses.

Weaknesses:

No weaknesses were identified by this reviewer.

Author response:

The following is the authors’ response to the original reviews.

eLife Assessment:

This important study investigates how distinct honey bee viruses differentially alter flight performance through interactions with octopamine signaling pathways. The combination of behavioral flight assays, pharmacological perturbation, and transcriptomic analyses provides solid evidence that virus-specific effects on flight are associated with octopamine signaling. However, some of the stronger mechanistic conclusions regarding direct regulation of octopamine signaling remain incomplete without more specific validation of receptor-level effects and direct quantification of octopamine levels or signaling activity.

We revised some of text in the manuscript, since we agree that octopamine and tyramine quantification would strengthen the mechanistic interpretation of our findings. While we acknowledge that direct measurements of OA and tyramine would provide valuable complementary evidence, the current study relies on multiple independent lines of evidence—including gene expression analyses, OA supplementation experiments, and behavioral measurements—that collectively support a role for octopaminergic signaling in mediating the observed effects. The revised text better reflects the data included in this paper.

Public Reviews:

Reviewer #1 (Public review):

Summary:

Kaku and Flenniken investigate the mechanistic pathways through which specific viral infections alter the flight capabilities of honey bees. Building on their previous discovery that DWV impairs flight while SBV unexpectedly enhances it, the authors hypothesized that these behavioral shifts are driven by interactions with the insect's octopamine (OA) signaling pathway, which is responsible for the "fight-or-flight" neurohormonal stress response and energy mobilization. To test this, the authors experimentally infected adult honey bees with DWV or SBV and pharmacologically manipulated the OA pathway using either octopamine supplementation or epinastine (EP), an OA-receptor antagonist. They then evaluated the bees' flight performance (distance, duration, and speed) on custom flight mills and profiled their gene expression using qPCR and RNA sequencing.

Strengths:

A major strength of this study is the high prevalence of preexisting background DWV and SBV infections in the honey bee cohorts, which meant there were no completely "virus-free" control groups. However, the authors successfully mitigated this limitation by rigorously quantifying viral RNA copies for every individual bee via qPCR and utilizing these viral abundances as continuous variables in powerful linear mixed-effect models.

Weaknesses:

The primary weakness lies in the methodology used for targeted pharmacological manipulations, as well as the lack of OA quantification across different treatments. Thus, their claims are not sufficiently supported by the current data.

We thank Reviewer #1 for these comments.

(1) The authors utilize Epinastine to block octopamine signaling, describing it as a highly specific OA receptor antagonist. However, pharmacological inhibitors often lack absolute specificity. Epinastine might bind to other octopamine receptor subtypes present in honey bee neural and flight muscle tissues, or it could potentially cross-react with tyramine and dopamine receptors. Without further genetic validation (e.g., RNA interference targeting specific receptors), it is difficult to definitively conclude that the altered flight performance is solely due to the blockade of the specific Oβ−2R pathway.

We thank the reviewer for this thoughtful comment and agree that pharmacological approaches have inherent limitations with respect to receptor specificity. However, among the available octopamine receptor antagonists, epinastine is considered one of the most selective compounds for insect octopamine receptors. Roeder et al. (1998) reported that epinastine exhibits affinities for octopamine receptors that are at least four orders of magnitude greater than those for other insect biogenic amine receptors, including dopamine, tyramine, histamine, and serotonin receptors. We updated the text to include this information.

Honey bees encode four β-adrenergic-like receptors (AmOARβ1- AmOARβ4) and one αadrenergic-like receptor (AmOARα1). Our transcriptomic analyses indicated that expression of AmOARβ2 was substantially higher than that of other octopamine receptor genes. Specifically, AmOARβ4 transcripts were not detected in our RNA-seq datasets, while AmOARβ1 and AmOARβ3 were expressed at very low levels in most samples (Supplementary Table S9; Figure S5). Although AmOARα1 transcripts were detected in some samples, expression levels were consistently lower than those of AmOARβ2. These observations support the interpretation that the physiological effects observed following epinastine treatment are primarily mediated through disruption of AmOARβ2 signaling. We updated the text to include this information.

We agree that receptor-specific genetic approaches would provide valuable complementary evidence. RNAi-mediated knockdown of AmOARβ2 is an attractive future direction; however, RNAi efficacy in honey bees is variable and influenced by factors including transcript turnover rates. In addition, dsRNA treatments can induce sequence-independent antiviral effects that could confound interpretation in studies involving viral infection (Flenniken and Andino PONE 2013; Brutscher, Daughenbaugh, and Flenniken Sci Reports 2017). We have revised the manuscript to more explicitly acknowledge these limitations and to clarify the basis for our interpretation of the epinastine experiments.

(2) As a natural neurotransmitter, insects have evolved highly efficient "cleanup" mechanisms. OA is rapidly cleared from the synaptic cleft via reuptake transporters and quickly inactivated by enzymes such as N-acetyltransferase (NAT) or Monoamine Oxidase (MAO). Consequently, an injection of OA produces only a transient "pulse" of activity. It is often a poor "tool" for inducing prolonged physiological effects compared to synthetic formamidines like Amitraz.

We thank the reviewer for this important point regarding the pharmacokinetics of octopamine. We agree that octopamine is rapidly metabolised and cleared under physiological conditions and that exogenous administration is unlikely to precisely mimic endogenous signaling dynamics. Our goal was not to induce a prolonged pharmacological activation of octopamine signaling comparable to that produced by synthetic agonists such as amitraz, but rather to determine whether increasing octopaminergic signaling could mitigate the flight impairments associated with DWV infection. Octopamine was administered either by injection or through feeding (Lines 86-89), both of which resulted in significant improvements in flight performance in DWV-infected bees (Figure 2). The observation that two independent delivery methods produced similar outcomes supports the conclusion that enhanced octopaminergic signaling can partially rescue the DWV-associated flight phenotype. We have revised the manuscript to clarify this distinction and to acknowledge that exogenous octopamine administration likely produces transient elevations in signaling rather than sustained receptor activation.

(3) The study relies heavily on transcriptomics and quantitative PCR to measure the mRNA expression of key synthesizing enzymes, namely tyrosine decarboxylase (tdc) and tyramine βhydroxylase (tβh), to infer the activation or suppression of the octopamine pathway. However, changes in enzyme synthesis at the RNA level are often insufficient to accurately reflect the true physiological levels of biogenic amines. To robustly prove the authors' hypothesis of a "feedback loop that regulates intracellular OA concentrations", direct quantification of actual octopamine and tyramine titers in the bees (e.g., using high-performance liquid chromatography or mass spectrometry) is necessary.

We thank the reviewer for this comment and agree that octopamine and tyramine quantification would strengthen the mechanistic interpretation of our findings. Previous studies have successfully quantified OA in honey bees using HPLC-based approaches, including KayaZee et al. (2022, eLife), who measured OA in honey bee muscle tissue (both naturally occurring levels and levels post-treatment with 10 mM OA), and Cook et al. (2017, J. Exp. Bio) who quantified OA in pooled honey bee brain samples.

Prior to submission, we inquired with our institutional mass spectrometry facility regarding the feasibility of measuring OA in individual honey bee samples. The expected concentrations of OA in our samples was below their limit of detection, so we did not pursue these analyses at that time. During the review process, we explored the possibility of analyzing a subset of samples at external facilities that may have the sensitivity required to quantify OA and tyramine in honey bee tissues. Since such analyses would require substantial resources, with estimated costs of approximately $5,000–10,000 for 12–15 samples that have been stored in the -80C since the study, rather than flash-frozen in liquid nitrogen as described by Zee et al. 2022. While we acknowledge that direct measurements of OA and tyramine would provide valuable complementary evidence, the current study relies on multiple independent lines of evidence— including gene expression analyses, OA supplementation experiments, and behavioral measurements—that collectively support a role for octopaminergic signaling in mediating the observed effects. We thank the reviewer for this valuable suggestion. While these analyses are beyond the scope of this study, we will consider using this approach in future studies.

Reviewer #2 (Public review):

Summary:

This highly original and well-designed study provides insight into how honeybee picorna-like viruses, Deformed wing virus (DWV) and Sacbrood virus (SBV), affect flight performance, and reveals the role of the octopamine (OA) pathway in virus-honeybee interactions. The authors used a flight mill to quantify the flight performance of bees with different levels of DWV and SBV. Bees were treated with OA and/or epinastine (EP) - an OA receptor antagonist; the study also quantified virus loads and expression of two key genes involved in OA biosynthesis.

The results showed that reduced flight performance associated with high DWV levels could be alleviated by OA administration. In contrast, increased levels of SBV had the opposite effect, leading to enhanced flight performance. This suggests distinct physiological responses to DWV and SBV infections. Administration of EP had led to a reduction of flight performance in SBVinfected bees, indicating the involvement of the OA pathway.

The authors also quantified levels of mRNAs of enzymes involved in OA synthesis, tyrosine decarboxylase (TDC) and tyramine beta-hydroxylase (TbH), and concluded that DWV induced expression of TbH, while SBV upregulated expression of TDC. Furthermore, the study identified upregulated and downregulated genes in response to SBV, DWV and DWV in combination with OA.

Strengths:

The study reported opposing effects of infections of related viruses, SBV and DWV, on honeybee flight performance, and identified the central role of the octopamine (OA) signaling pathway in the effect of viruses on honeybee flights.

These findings were achieved by using a combination of approaches, including experimental measurement of flight distance, virus infections, and introduction of OA and EP. Experimental work with honeybees is technically challenging and requires specialized expertise, which makes the results produced in this study more valuable.

DWV and SBV are among the most important honeybee pathogens affecting honeybee health and threatening the pollination service. Therefore, an understanding of the mechanisms underlying DWV and SBV pathogenesis has the potential to develop novel approaches to mitigate the negative impact of these viruses.

Weaknesses:

No weaknesses were identified by this reviewer.

We thank Reviewer #2 for these comments

Recommendations for the authors:

Reviewer #2 (Recommendations for the authors):

I have only minor suggestions for the manuscript.

(1) L. 45-46

Please note that not only high virus levels have a negative impact on honeybees. Low levels of DWV, typical of covert infections, can have long-term deleterious effects on honeybee foraging and survival. Please include citation (e.g., Benaets et al, 2017, Proc Biol Sci (2017) 284 (1848): 20162149. https://doi.org/10.1098/rspb.2016.2149).

We thank the reviewer for these comments and edited the text accordingly, and apologize for our inadvertent omission of Benaets et al 2017, which is cited in our previous publication.

(2) L. 113

Clarify what is meant by "high DWV levels"

"i.e., 10^8 copies / 2 ug RNA" -> "i.e., above 10^8 copies / 2 ug RNA"

We thank the reviewer for this comment and corrected this in the text.

(3) L.115

"..mock infected bees.." /Figure 2A.

Did these bees have low levels of DWV, below 10^8 / 2 mg RNA? What was the level of DWV in these bees?

Mock-infected bees had an average of 3x105 DWV copies and 2x103 SBV copies per 2 µg RNA (reported in Lines 107-109 in revised manuscript, a few lines before in original manuscript).

(4) Figure 2 / Legends to Figure 2

Note that in Figure 2 legends, the grey areas show 95% confidence intervals for regression lines.

We thank the reviewer for this comment and added this in the text.

(5) Figure 2 / Legends to Figure 2

Consider including correlation coefficients (R) and p-values for each of the regression lines in Figures 2A-F. (These could be included in the Figure 2 legends).

We thank the reviewer for this suggestion and agree that providing sufficient statistical information is important for data interpretation. Because the analyses presented in Figure 2 are based on linear mixed-effects models that incorporate both fixed and random effects, the statistical outputs are more complex than those associated with simple linear regressions. For figure clarity, we chose not to include all model statistics within the figure panels or legends and the key statistical results, including p-values and model fit metrics (R2 values), are reported in the main text (Lines 121+). In addition, complete model outputs, including all relevant coefficients, correlation estimates, and associated statistics, are provided in Supplemental Data Sheet S4. To address the Reviewer’s comments, we revised the figure caption to improve clarity and include key p-values.

We believe this approach better balances accessibility in the main figures with comprehensive reporting of the statistical analyses and thank the Reviewer for this useful suggestion.

(7) L.357-377 - virus-specific responses

A previous honeybee transcriptome analysis study, which showed different responses to DWV and SBV, could be cited (Ryabov E. 2016. PeerJ 4:e1591 https://doi.org/10.7717/peerj.1591).

We thank the reviewer for this point and included this citation in line 338 of original manuscript (line 349 in revised, tracked-changes manuscript).

(7) L. 412

"bees were collected 24 hours prior to eclosion" -> e.g. "bees were collected at pupal stage 24 hours prior to eclosion"?

Specify if dark-eyed pupae were collected to make sure eclosion in 24 hr.

We thank the reviewer for making this point, and we revised the methods and results text to improve clarity.

Version 2

Abstract

High annual honey bee colony losses are associated with environmental and biological stressors, including virus infections. In insects, the octopamine pathway orchestrates the “fight-or-flight” response, regulating energy mobilization, temperature, and flight. We determined that sacbrood virus (SBV) infections induce expression of an octopamine receptor and enhance honey flight performance, whereas deformed wing virus (DWV) infections reduce flight performance, but how viruses interface with this pathway remained unknown. To elucidate the relationships between the octopamine response, virus infection, and flight, honey bees were infected with SBV or DWV and exposed to octopamine (OA), epinastine (EP)–an OA receptor antagonist, or both OA and EP; flight and gene expression were assessed. Pharmacologic manipulation revealed that octopamine supplementation rescued flight deficits in DWV-infected bees, but diminished performance in SBV-infected bees, while blocking octopamine receptors altered these effects. Transcriptome analyses indicated that SBV infections, and DWV infection with OA treatment, activated honey bee metabolic pathways, and that SBV infected bees had greater expression of genes involved in OA synthesis, unless treated with OA. These results provide mechanistic insight for virus-specific impacts on honey bee flight, which may have consequences on foraging efficiency, colony health and virus transmission.

Introduction

Honey bees (Apis mellifera) are eusocial, generalist pollinators that live in densely populated colonies of ∼30,000 individuals (1–3). The survival of each colony depends on the coordinated activities of thousands of individual bees that provision nutritional resources, rear brood, and regulate internal hive temperature (3–6). Colony losses in the U.S. averaged 40% from 2008-2025 (7–11), these losses are associated with several abiotic and biotic stressors, including parasites and pathogens (12–16). Honey bee pathogens include bacteria, microsporidia, and viruses (17, 18). Viruses may be transmitted vertically, horizontally, or by vectors, including the Varroa destructor mite (19–23). Varroa mites feed on honey bees and infestations may kill colonies (24). In addition, Varroa mites vector viruses, including deformed wing virus (DWV, Iflavirus aladeformis), which replicates in the mite, and others that are mechanically transmitted such as sacbrood virus (SBV, Iflavirus sacbroodi) (25–27). Most characterized bee infecting viruses have positive single-stranded RNA (+ssRNA) genomes including members of Iflaviridae (e.g., DWV and SBV). Virus infections in honey bees may remain seemingly asymptomatic, cause morphological symptoms, paralysis, and/or result in death (17, 28–32). Honey bees that lack morphological symptoms may still harbor virus infections that impact lifespan and flight performance (8, 33–39).

Flight is an energetically taxing behavior that is essential for honey bee foraging, defecation, and mating (40–45), and flight muscles are essential for temperature regulation (i.e., fanning to cool and ‘shivering’ for heat) (19, 46–49). In our previous work, we used flight as a proxy for overall bee health and demonstrated that honey bees harboring high levels of DWV without obvious symptoms had impaired flight performance (33). Unexpectedly, we determined that the flight performance of SBV infected honey bees was enhanced and linked to the metabolism-stimulating, stress-induced octopamine pathway (33). Abiotic and biotic stressors may induce enhanced activity of stress pathways including the octopamine response. Invertebrates synthesize octopamine (OA), a ‘fight or flight’ neurohormone analogous to norepinephrine, which facilitates energy production for taxing activities including flight (50–54). The octopamine response may also be triggered by other stressors, including pathogen infections (33, 55, 56).

Whether octopamine signaling is immunosuppressive or immunoenhancing is dependent on multiple factors and its role in honey bee antiviral defense has not been thoroughly investigated (57, 58). Honey bee behaviors may be influenced by multiple monoamine neurohormones including OA, serotonin, dopamine, and tyramine (51, 59–64). Octopamine promotes movement, neuromuscular signaling, and metabolism. To synthesize OA, tyrosine is first converted to tyramine by Tyrosine decarboxylase (TDC); tyramine is the precursor neurohormone to OA and is generally associated with reduced flight and movement (52, 65, 66). Tyramine may serve as an antagonistic modulator of behavior to OA or Tyramine β-hydroxylase (TβH) may convert it to OA (Fig 1A). One of the most abundant OA receptors in honey bees is a G-coupled protein receptor, Oβ-2R (67). When OA binds to Oβ-2R, it activates adenylyl cyclase (AC), which increases intracellular concentration of cyclic adenylyl monophosphate (cAMP). Elevated cAMP activates cAMP-dependent Protein kinase A (PKA) which phosphorylates metabolic enzymes and transcription factors that generally enhance metabolic activity (Fig 1B) (47, 53, 67, 68).

Octopamine signaling pathway (i.e., synthesis, activation, inhibition) and experimental design

(A) Octopamine (OA) is synthesized by conversion of tyrosine to tyramine by tyrosine decarboxylase (TDC) and tyramine conversion to OA by tyramine β-hydroxylase (TβH). (B) OA β-2 receptor (Oβ-2R) is one of the most abundant and specific OA-detecting receptors. Upon activation of Oβ-2R, a G-coupled protein receptor, adenylyl cyclase (AC) is activated and increases intracellular cyclic adenylyl monophosphate (cAMP) levels. High cAMP levels induce protein kinase A activity which phosphorylates metabolic enzymes and transcription factors that enhance metabolic activity. (C) An OA receptor antagonist, epinastine (EP), inhibits OA binding and prevents activation of the Oβ-2R pathway. (D) To evaluate relationships between virus infection and the OA response, honey bees were inoculated with either deformed wing virus (DWV) or sacbrood virus (SBV) via injection or mock infected. Bees were exposed to either OA, EP, or both OA and EP via co-injection or feeding, all bees were fed sucrose. (E) Impacts of experimental conditions were quantified in individual bees 72 hours post infection. Panels B-D were produced using BioRender.com.

We hypothesized that SBV infections, which correlate with higher expression of an OA receptor (Oβ-2R) (33), activate the OA response which results in enhanced flight performance. Therefore, we predicted that OA treatment would enhance honey bee flight performance regardless of virus infection. Epinastine (EP) is a highly specific OA receptor antagonist, exhibiting at least four orders of magnitude greater affinity for OA receptors than for other honey bee biogenic amine receptors (69, 70). Therefore, we hypothesized that EP treatment would reduce honey bee flight performance and negate SBV-associated enhanced flight performance.

Results

To test these hypotheses, we exposed virus infected honey bees to OA, EP, or both OA and EP and assessed flight performance and gene expression (e.g., tdc, tβh, and oβ-2R) in three independent experiments (i.e., exp-1, 2, and 3) (Fig 1 and Data S1). Specifically, age-matched, one day old adult bees (n = 336) were experimentally infected with DWV (n = 117), SBV (n = 75), or mock infected (buffer injection, n = 144) and treated with OA (10 mM), EP (4 mM), or both OA and EP via injection or feeding, or not treated; all bees were fed 50% sucrose syrup ad libitum. Treatments of OA and EP were given in 2 µL injections or fed at the same concentration in sucrose syrup; doses were in-line with previous studies (33, 47, 52, 69). Exposure to OA and/or EP via injection was temporally limited, whereas ad libitum feeding provided continuous exposure (Fig 1).

Quantification of experimentally introduced and naturally occurring honey bee virus infections

Honey bees used in this study were sourced from managed colonies that are naturally exposed to pathogens. To limit variation associated with confounding infections, bees were collected upon emergence under laboratory conditions (19, 33, 71, 72). Therefore, in addition to the use of quantitative polymerase chain reactions (qPCR) to determine the abundance of DWV and SBV in individual honey bee samples, PCR was used to test for potential preexisting infections and detected pathogens were quantified using qPCR, as described previously (33) (Data S1-S3). We determined that honey bees in exp-1 had preexisting DWV infections and bees in exp-2 and exp-3 had preexisting infections of DWV and SBV (S1 Fig). Most bees analyzed herein had either experimentally introduced or preexisting DWV infections (n = 325/336) and 40% had experimentally introduced or preexisting SBV infections (n = 134/336). DWV infections ranged between 1×103 and 7.1×1010 copies per 2 µg of RNA (∼7×103 – 8×1011 DWV copies per bee).

SBV infections were between 1.2×103 and 1.2×1010 copies per 2 µg of RNA (∼8.3×104 – 8.3×1011 SBV copies per bee). Due to high DWV and SBV prevalence there were no virus-free honey bees. Mock infected bees harbored low virus levels (i.e., an average of 3×105 DWV copies and 2×103 SBV copies per 2 µg RNA) and inoculation resulted in higher virus loads (i.e., an average of 2×108 DWV copies and 2×108 SBV copies per 2 µg RNA), p = 2×10−16 and 5×10−11, respectively). To compare the impact of varying virus levels and treatments, we evaluated several flight metrics (i.e., flight distance, duration, speed, and stop count; Data S1) using linear mixed effect models (LMMs) (complete information for each model and model selections in methods and Data S4).

OA and EP altered flight performance of virus infected bees

The previously described DWV-associated flight impairment relative to mock infected bees (33), was recapitulated in this study (i.e., DWV infected bees flew shorter distances (p = 0.005)). Mock infected bees from this sample cohort, which harbored low virus levels, flew an estimated 45 meters, whereas bees harboring high DWV levels (i.e., > 108 DWV RNA copies / 2 µg RNA) flew only ∼9 meters. As expected, DWV-associated flight impairment was alleviated by OA feeding or injection (i.e., average flight distance was similar to mock infected bees fed sucrose, p = 0.61 and 0.08, respectively) (Fig 2A). Honey bees that received EP flew shorter distances (p = 0.002), but the effect of feeding EP was not assessable since bees with experimentally introduced DWV infections that were fed EP died, and there were only 14 bees with low DWV infection levels (Fig 2B, Data S4). Co-injection of OA and EP negatively impacted the distance DWV infected honey bees flew (p = 0.020) (Fig 2C). Counter to our hypothesis, OA-treatment did not enhance flight performance of SBV infected bees (p = 0.69 and 0.60) (Fig 2D, Data S4). SBV infected bees that were treated with EP or co-treated with both EP and OA via injection flew shorter distances than untreated SBV infected bees (p = 0.022 and 0.004, respectively), whereas feeding had no impact (Fig 2E-F).

The flight performance of honey bees infected with DWV and/or SBV is differentially impacted by OA and EP treatment

To evaluate the involvement of the OA pathway on virus infection and/or flight performance, virus- or mock- infected honey bees were exposed to OA, EP, or both OA and EP via co-injection (blue) or feeding (green). Honey bees were experimentally infected with or had preexisting infections of DWV (A-C) or SBV (D-F) in three experiments. Flight distance (log10 m) was evaluated using linear mixed effect models to quantify impacts of DWV and/or SBV infection in the context of OA and/or EP treatment. Each data point represents an individual bee (n = 336); colors indicate treatment (i.e., bees fed only sucrose in pink, bees fed OA and/or EP in green, and bees injected with OA and/or EP in blue) and gray areas represent 95% confidence intervals. Linear mixed models with estimates in Data S4. (A) In bees fed only sucrose, DWV infections negatively impacted flight distance (p = 0.005), whereas DWV infected bees treated with OA flew further distances (i.e., average flight distance was similar to mock infected bees fed sucrose, p = 0.61 and 0.08, respectively). (B) DWV infected bees injected with EP flew shorter distances than DWV infected bees fed only sucrose (p = 0.002). (C) Co-injection of OA and EP negatively impacted the distance DWV infected honey bees flew (p = 0.020) (D) OA-treatment did not enhance flight performance of SBV infected bees (p = 0.69 and 0.60) (E) Honey bees with SBV infections fed or injected with EP flew shorter distances than SBV infected bees fed only sucrose (p = 0.022 and 0.004, respectively). (F) Bees with SBV infections fed or injected with OA and EP resulted in similar flight distances regardless of SBV infection level.

The amount of time a honey bee was actively flying was also assessed as an indicator of energetic capacity (S2 Fig). Honey bees harboring high DWV levels (i.e., >108 DWV copies / 2 µg RNA) flew only 14 minutes (p = 0.022), relative to mock infected bees with low virus infection levels (average of 3×105 DWV copies / 2 µg RNA), which flew for 26 minutes. Honey bees with high SBV levels flew for similar durations to mock infected bees (i.e., 28 minutes, p = 0.58), which is consistent with previous results (33). In line with our hypotheses, mock infected bees with low virus infection levels flew longer when they were fed OA (i.e., 49 minutes) (p = 0.0001). DWV-associated flight impairment was mitigated in honey bees that were supplemented with OA, which flew for similar durations to mock infected bees (i.e., 28 minutes, p = 0.76). However, counter to our hypothesis, when SBV infected bees were injected or fed OA, flight duration was significantly reduced (i.e., 18 and 12 minutes, respectively; p = 0.04 and 0.0009).

To quantify the impact of virus and OA and/or EP treatment on honey bee flight speed, we compared average and peak speeds. Average flight speed (i.e., total distance / total flight duration) was similar across all conditions except in bees fed EP, which flew at slower average speeds (p = 3×10−4) (Data S4). We also compared peak speed (i.e., maximum speed reached during flight sustained >10 seconds) (S2 Fig). Similar to previously reported results (33), DWV infected bees had slower peak speeds than mock infected bees (p = 0.016). SBV infected bees flew similar speeds to mock infected bees (p = 0.07), but peak speeds were reduced when SBV infected bees were fed or injected with OA (p = 0.004, p = 0.0006) (Data S4). Mock infected bees flew at peak speeds of ∼3.3 km/h, honey bees with high levels of SBV (i.e., 108 SBV copies / 2 µg RNA) flew ∼3.8 km/h, whereas highly SBV infected bees injected or fed OA flew an estimated 2.0 or 2.5 km/h, respectively (p = 0.001 and 0.022).

In addition to flight metrics, post-flight thorax temperatures were measured within a subset of individual bees (n = 130) since flight induces heat and the octopamine response is involved in honey bee temperature regulation (47, 67) (Data S4). The ambient temperature for all flight assays was 22°C, and post-flight thorax temperatures in honey bees maintained in control conditions averaged 28°C. DWV infected bees had similar post-flight temperatures, and highly SBV infected bees were 2°C cooler than control (p = 0.04). Honey bees with DWV infections that were fed OA had thorax temperatures that were 2.5°C higher than bees not fed OA. Whereas the thorax temperatures of bees with DWV infections were 5.6°C lower in bees co-fed OA and EP (p = 0.027 and 1×10−6). Likewise, honey bees with high SBV levels that were injected with OA also had reduced thorax temperatures post-flight (i.e., ∼ 3°C lower) (p = 0.03; Data S4).

Virus specific outcomes of experimentally manipulated octopamine pathway explained by differential expression of key genes

Induction of the OA pathway was previously implicated in the enhanced flight performance of SBV infected honey bees, which had elevated expression of oβ-2R (33). When OA binds to Oβ-2R, it activates AC, increases cAMP concentration, thereby activating cAMP-dependent PKA, which phosphorylates enzymes and transcription factors that enhance metabolic activity (Fig 1B) (47, 53, 67, 68). To determine if DWV, SBV, or different experimental treatments were associated with differences in OA synthesis, we quantified the expression of tdc and tβh relative to mock infected honey bees fed only sucrose (Data S1, Fig 3). We evaluated the relationship between tdc expression and multiple factors including DWV abundance, SBV abundance, experimental treatments (i.e., OA and/or EP, fed/injection) using LMMs which explained 91% of the data variation (Data S4). SBV infection was associated with increased tdc expression (p = 0.021), and SBV infected bees fed OA or co-fed OA and EP also expressed higher tdc levels (p = 0.030 and 2×10−10, respectively). Expression of tdc was lower in SBV infected bees co-injected with EP, and SBV and DWV-coinfected bees that were fed OA or both OA and EP (p-values = 0.033, 0.022, 0.006, 0.004, respectively) (Fig 3A). Expression of tdc was not associated with DWV infection, or DWV infection with OA and/or EP treatment (p-values all >0.2, Data S4, S3 Fig). We also examined the effects of virus infection in the context of experimental OA and EP treatments on tβh expression (i.e., in bees with detectable tβh levels n=306/336) (Fig 3B). An LMM that included DWV abundance, SBV abundance, treatment, and tdc expression as explanatory variables explained 85% of the variation in tβh expression (Data S4). Expression of tdc and tβh were positively correlated (with a tdc:tβh ratio of 1:1.6, p = 0.004) (S4 Fig). Expression of tβh was higher in SBV and DWV infected bees (p =0.022 and 1×10−9, respectively), suggesting more tyramine is converted to OA during DWV or SBV infection (Fig 3B). Results from these models indicate that only SBV infections were associated with greater expression of both tdc and tβh relative to mock infected bees with low preexisting virus levels. Whereas SBV infected honey bees that were co-injected with OA expressed similar levels of tdc or tβh as mock infected bees (p = 0.421 and 0.437, respectively). This result suggests that the elevated expression of genes involved in OA synthesis in SBV infected honey bees is reduced by OA supplementation and partially explains the impaired flight performance of these bees relative to untreated SBV infected honey bees (Fig 4). For OA to stimulate metabolic activity and energy production, OA must activate receptors including Oβ-2R (67). To identify variables that contribute to the range of oβ-2R expression, we compared LMMs including tβh expression, tdc expression, SBV abundance, and OA/EP treatment, which explained 95% of the data variation (S4 Fig). We considered including DWV abundance as a fixed effect in this model, but it did not appreciably contribute to oβ-2R expression (Data S4). These results corroborate our previous findings that SBV, but not DWV, was associated with increased oβ-2R expression (33). The strongest predictors of oβ-2R expression levels were tdc and tβh expression (p = 4.4×10−5 and 3.9 × 10−5) and SBV was positively associated with tdc and tβh expression (p = 0.0006) (Data S4).

Expression of tyrosine decarboxylase (tdc) and tyramine β-hydroxylase (tβh) across treatments and SBV infection

The relationships between honey bee tyrosine decarboxylase (tdc) and tyramine β-hydroxylase (tβh) expression in the context of virus infection and treatments were evaluated using a linear mixed effect model that included SBV abundance and OA/EP treatments as fixed effects with an interaction. Individual honey bee data plotted by treatment group with either (A) tdc expression or (B) tβh expression. Gene expression was calculated as log2 fold change using the ΔΔCt method relative to housekeeping rpl8 and compared to mock infected bees. SBV abundance was calculated as log10 copies / 2 µg RNA. The background violin plots represent the 95% confidence interval, and the horizontal line in each violin represents the median. SBV abundance is represented by a green color scale and SBV-negative samples with an “X”. When SBV abundance did not impact tdc or tβh expression the dark green points are evenly distributed above and below the control ‘0’ fold change line, whereas unequal distribution indicates an SBV-specific effect; * p<0.05, **p<0.01, and ***p<0.001. (A) Positive association between SBV abundance and tdc expression (p = 0.021). SBV infected bees fed OA or both OA and EP had greater tdc expression levels than uninfected bees fed only sucrose (p = 0.03 and 2×10−10). (B) Expression of tβh was greater in SBV infected bees than uninfected bees (p =0.022). SBV infected bees fed OA had greater tβh expression levels than uninfected bees fed sucrose (p = 2×10−5). SBV infected bees injected with EP or co-injected with OA and EP had lower tβh expression (p = 2×10−5 and 4×10−9).

Relative tdc and tβh expression in honey bees with high DWV or SBV levels

(A) Tyrosine is converted via tyrosine decarboxylase (TDC) to tyramine. Tyramine is a neurohormone that acts as a behavioral antagonist to octopamine (OA), resulting in reduced movement in invertebrates. Tyramine may act as a signaling neurohormone, or tyramine β-hydroxylase (TβH) may convert it to OA. (B) A linear mixed model was used to assess the relative expression of tdc and tβh in honey bees harboring high levels of either deformed wing virus (DWV) or sacbrood virus (SBV) (i.e., 108 virus copies / 2µg RNA) in the context of OA and/or epinastine (EP) injection. Estimates of relative tdc and tβh expression were generated based on data from 336 honey bees. The single bold dash indicates tdc or tβh expression was similar in virus infected bees and mock infected bees and arrows indicate estimates that were higher or lower than controls where each arrow represents fold change of 1.

Transcriptome level impacts of virus infection, flight, and OA treatment

We hypothesized that flight-impacting virus infections would result in differential expression of genes involved in stress, metabolic, and immune pathways. To test this, the transcriptomes of SBV, DWV, DWV and OA, or mock infected honey bees were compared. Honey bees were obtained from a previous experiment and included individuals that did or did not fly (33). RNA was prepared for high throughput sequencing on an Illumina NovaSeq X Plus 25B (n = 36 total individual bee samples, 3-5 per treatment group; Data S6). An average of 25 million reads was obtained for each sequencing library and mapped to the A. mellifera assembly HAv3.1 (73, 74). To evaluate similarities and differences between the transcriptional responses of honey bees with flight-impacting virus infections, we compared DWV, SBV, and DWV and OA injected bees relative to mock infected bees immediately post-flight since these bees had little to no detectable preexisting infections, whereas the no-flight mock infected bees had preexisting DWV infections (Data S5-S6).

Hundreds of differentially expressed genes (DEGs) were identified in SBV, DWV, and OA treated DWV infected bees (Fig 5). DWV infected bees with and without OA treatment shared 95 DEGs, which was more than they shared with SBV infected bees, indicating that there are virus specific transcriptional responses. There were 24 genes shared among SBV, DWV, and DWV and OA injected bees including elevated expression of those involved in the Ras-Raf-MEK-ERK pathway (including mesr3 and rap2l) which is associated with metabolism, muscle development, and the cAMP/PKA pathway (75). Genes with lower expression were involved in muscle function (i.e., mlc2, mlc1, cher, sar), energy production (i.e., cnn, ant, pka-c3), and an inhibitor of proteolytic activity (i.e., cyp4g11).

Transcriptome level comparison of honey bees that were SBV infected, DWV infected, or DWV infected and OA-treated

Gene expression was evaluated in honey bees that were mock infected, SBV infected, DWV infected, or DWV infected bees that were co-injected with OA (DWV+OA) after flight. Differentially expressed genes (DEGs) for each treatment group were identified relative to mock infected bees and compared using a Venn diagram. Full list of DEGs, fold change, and Benjamini-Hochberg corrected significance values are included in SI Table S6.

There were many genes unique to SBV and DWV infections (Fig 5). Honey bees with SBV infections exhibited lower expression of several genes associated with the citric acid cycle (acsl6, cg5065, acyl-coA delta(11), mcad, nd-13a, nd-13b, nc73ef, and mdh2) and general muscle function and tissue repair genes (tmp-1, tmod, plastin-2, and fgf17). SBV infected bees had elevated expression of rgs11–which increases G-protein α-subunit active state to inactive to expedite the G-coupled protein response (76), titin–which contracts stretched/striated muscle tissues (77), and several immune genes such as argonaute-2, relish, and domeless (Data S6-S7). DWV infected bees exhibited greater expression of several genes associated with immune function including protein mahjong, spastin, and a small heat shock protein, protein-lethal(2) essential for life-like (Data S6). The DEGs unique to DWV infected bees post-flight that exhibited lower expression included genes involved with muscle function and structure (obscurin, a titin homolog, ryr, actin, mhc, stacl, tpm1, and zasp66) and ATP generation (citrate synthase, adenylate kinase 1, ADP/ATP translocase).

Since OA supplementation mitigated DWV-associated flight impairment, we identified differences associated with OA by comparing DEGs to untreated DWV infected honey bees post-flight. As hypothesized, DWV infected bees that were co-injected with OA treatment had greater expression of OA pathway genes (oβ-2R, adenylyl cyclase (ac), and CREB binding proteins) and genes involved in glucose metabolism (gale, gdh, adpgk, putative glucosylceramidase 4, gba1lp, g6pd, sugarless, pdp, ND-24, NAD kinase, tret1, 1,5-anhydro-D-fructose reductase, and igfals) (Data S8-S9). These bees also expressed higher levels of genes induced by high cyclic guanine monophosphate (cGMP) levels (i.e., burgundy, cGMP-dependent protein kinase-like, and cGMP-dependent protein kinase 1) and genes in the RAS/MAPK-pathway (rho1, rac1, and rab5). Untreated, DWV infected bees expressed higher levels of several immune genes (abaecin, domeless, echinoid, and defensin-1) and heat shock protein encoding genes (hsp83, hsp90, hspβ-1) relative to DWV infected bees treated with OA (Data S7-S9).

After flight, both SBV infected and OA treated DWV infected bees expressed lower levels of several mitochondrial associated genes (vdac, blw, atpsynβ, cox10, klf6, and mcmp1) and genes involved in energy generation (pyruvate kinase, gapdh2, and 2 ND-49, and ND-51) than mock infected bees (Fig 5). This result suggested that flight was more taxing for virus infected bees. In addition, transcriptome level results indicated that OA treatment during DWV inoculation resulted in enhanced energy production, since these bees expressed higher levels of genes involved in energy production post-flight relative to DWV infected bees that did not fly (Data S10).

Since OA treatment resulted in recovered flight performance in DWV infected bees, we compared the transcriptomes of honey bees with similar DWV-levels that were either OA-treated or untreated. The 50 genes with the greatest difference in mean expression were identified in bees that did or did not fly (i.e., 100 total possible genes) (Data S10). There were 30 shared genes associated with OA-treatment regardless of flight status. This list included defensin-1, apidermin-2, melittin, odorant binding protein 3, chymotrypsin-1, two α-glucosidases, peritrophin, 2 chitinase proteins, fibrillin-2, muscle protein 20, and 7 uncharacterized genes. The OA-associated list also included vegetative cell wallprotein-like (gpl-1), which exhibited higher expression in bees that flew further distances (mock, SBV, DWV and OA), than in DWV infected bees. Previous studies determined that gpl-1 expression was greater in Nosema ceranae infected honey bees (78), and that octopamine signaling was also impacted (55, 56). While the bees in our study were not Nosema infected (S1 Fig), we used qPCR to confirm that bees expressed higher gpl-1 levels when injected with OA or EP (LMM, p = 2×10−7 and 4×10−8, respectively); data from additional experiments (i.e., exp-1, n = 61 and exp-2, n = 69; Data S1). We also examined expression of a subset of OA-associated genes including OA receptors (i.e., oα-1R, oβ-1R, oβ-2R, oβ-3R, and oβ-4R), enzymes involved in OA synthesis (tdc and tβh), and genes associated with Oβ-2R activation (adenylyl cyclase (ac), protein kinase A (pka), and CREB-associated protein encoding genes) (Data S9 and S5 Fig). No sample had detectable levels of oβ-4R; oβ-1R and oβ-3R were expressed at very low levels in most samples (Supplementary Table S9; Figure S5). Although oα-1R transcripts were detected in some samples, expression levels were consistently lower than those of oβ-2R. Expression of tdc, tβh, oβ-1R, crebA, and creb-binding isoform X1 was generally higher in SBV infected honey bees, and greater tdc and tβh expression was confirmed via qPCR (exp 1-3, n = 336) (Fig 3, Data S1 and S4). Honey bees that were simultaneously DWV infected and treated with OA expressed higher levels of oβ-2R, oα-1R, and creb binding isoform X2 than untreated DWV infected bees (S5 Fig); greater oβ-2R expression was also confirmed by qPCR (LMM, n = 336; p = 9×10−4; Data S1). Expression of oa-1R and tyrR was assessed via qPCR in only one experiment since expression was generally low regardless of treatment (exp-3, n = 69; Data S1). SBV infected bees had greater expression of oa-1R (p = 0.017), and SBV infected bees co-injected with EP had significantly lower expression of both oa-1R and tyrR (p = 0.0006 and 0.04, respectively).

Discussion

The interactions between virus infections, host immune and stress responses, and organismal health are complex. Insects rely on OA as a neurohormone and neuromodulator to regulate physiological processes and behaviors (50, 52, 79–81), and our previous studies indicated that viruses differentially impacted honey bee octopamine signaling and flight performance. Therefore, we used honey bees as a model system to further investigate these interactions at organismal and transcriptional levels.

As previously described and shown herein, inapparent (or asymptomatic) DWV infections negatively impact honey bee flight performance (i.e., flight distance, duration, and peak flight speed). Evaluation of transcriptional responses to DWV infection suggest that DWV-associated flight impairment may be partially explained by lower expression of genes involved in energy production, which were elevated in OA treated honey bees (Fig 5, S5 Fig, Data S7-S8). These results are supported by previous studies that showed DWV infected honey bees expressed lower levels of G-coupled protein receptor encoding genes, as well as other genes involved in metabolism (82–84). Previously, we demonstrated that the flight performance of SBV infected honey bees was enhanced and that these bees expressed more oβ-2R (33). Herein, we showed that SBV-associated flight enhancement was reduced when OA signaling was inhibited by EP, an OA receptor antagonist (Fig 2 and 4). We expected that SBV infected bees would have compounding, enhanced flight performance when treated with OA, but unexpectedly, OA supplementation negatively impacted flight performance (Fig 2). To further investigate this physiological response, we examined the expression of key genes in the octopamine pathway and determined that in addition oβ-2R, SBV infected honey bees expressed greater levels of tdc and tβh (Data S1). Expression of tdc and tβh was lower in SBV infected bees that were treated with OA (Data S1 and S4, Fig 3-4). Reduced expression of these genes may limit OA synthesis, OA signaling, reduce metabolic activity, and result in less availfable energy for flight (50). Therefore, the reduced flight performance of SBV infected bees treated with OA may be explained by a feedback loop that regulates intracellular OA concentrations, though future studies that measure OA levels are needed to validate this hypothesis.

Depending on OA concentration and specific pathogen challenge, OA may enhance or inhibit immune function in many invertebrates, though mechanistic studies have been primarily restricted to two model organisms (57, 58, 85–87). In Drosophila, OA treatment impacts intracellular OA concentration, in a hypothesized feedback mechanism that regulates neurohormonal signaling (62, 88, 89). In Caenorhabditis elegans, OA pathways are hypothesized to be constitutively active, but downregulated upon pathogen infection to enable an effective immune response while limiting immune activity in the absence of pathogen challenge (85). Transcriptomic analysis of honey bees infected with a model virus (Sindbis) that were co-injected with double-stranded RNA, a virus-associated molecular pattern, had greater expression of an OA receptor (oβ-2R) (72). Results presented herein indicate that OA treatment of DWV infected honey bees alters expression of several genes associated with immune function including apidermin, defensin, and melittin (90–93). Additionally, DWV infected bees with OA treatment had reduced expression of an odorant binding protein that was previously associated with SBV infection, indicating that it may be involved in honey bee antiviral defense (94). Together, these results highlight virus-specific transcriptional responses in honey bees that may impact virus dynamics and flight performance during commonly occurring coinfections (33, 95, 96).

Multiple OA receptors have been identified in honey bee flight muscles and neural tissues including OαR1, Oβ-2R, Oβ-1R, and Oβ-3/4R (47). OA receptors have also been detected on invertebrate immune cells (57). In addition to their involvement in flight, OA receptors are also involved in honey bee hygienic, communicative dance, and thermoregulatory behaviors (47, 59, 63, 67, 97, 98). Typically, around 12 days post-emergence honey bees change from nurse bees, which care for developing brood, to pre-forager bees and undergo physiological changes required for flight (99–101). As honey bees age and become foragers, OA levels in the brain increase (51, 61, 102). Honey bees infected with the microsporidian parasite, Nosema, have greater OA levels than uninfected bees (55, 56) and are prone to precocious foraging (103), though Nosema levels did not appreciably impact honey bee flight capability (34). These findings are in-line with SBV-induced flight enhancement that correlated with higher oβ-2R expression levels (33). Previous laboratory-based studies demonstrated that precocious foraging may be induced by OA supplementation (51).

Honey bees are exposed to environmental OA sources including citrus blossom nectar, which may contain tyramine and OA at concentrations up to 256 µM (104–107). In addition, acaracides that are OA receptor agonists including amitraz and chlordimeform are commonly used in agriculture (108). Amitraz is commonly used to mitigate Varroa mite infestations in honey bee colonies (109). Varroa mites with amitraz-resistant Oβ-2R mutations have been reported (110, 111). Amitraz has generally been considered safe for honey bees since honey bee OA receptors are divergent from Varroa mite OA receptors (i.e., 48% identical at the amino acid level) (109, 112). However, honey bees exposed to amitraz had higher mortality when co-stressed via infection with a model virus (Flock House virus) (113). These results in conjunction with our findings, linking virus infection with the OA stress response pathway, may indicate previously unrecognized impacts at the organismal level. Varroa mites vector many honey bee viruses and infested colonies often have coinfections and greater average virus levels (114). Therefore, application of miticide to reduce Varroa and control vectored viruses, or the accumulation miticide metabolites in wax or other hive matrices, may inadvertently harm co-stressed, virus infected honey bees (113, 115, 116). Collectively, the data presented herein demonstrate that the OA pathway is a key factor in mediating virus-specific effects on honey bee flight performance, which may in turn influence viral transmission and colony health.

Materials and methods

Honey bees

Honey bees (Apis mellifera, primarily carnica), were maintained with standard apicultural practices at Montana State University’s Horticulture Farm. Each experiment was performed with honey bees from different colonies to obtain robust results from genetically diverse bees (Data S1). Frames with newly emerging bees were collected one day prior to each experiment and maintained at 32◦C in a laboratory incubator overnight. Young, age-matched (∼24 h post-emergence), female adult bees were utilized for all experiments. Bees were cohoused in modified deli containers (n = 10 per house), maintained at 32°C in an incubator and fed 50% sucrose syrup and water ad libitum (72). All honey bees were fed only sucrose syrup, unless octopamine (OA) and/or epinastine (EP) oral treatment was otherwise specified. OA was administered at a concentration of 10 mM and EP at 4 mM in fresh sucrose syrup daily, and honey bees were given unrestricted access to the supplemented sucrose syrup for the duration of the experiment.

Virus stock preparation

Deformed wing virus (DWV) and sacbrood virus (SBV) inoculums were propagated in white-eyed pupae as previously described (33). In brief, pupae were injected with 2 µL of Tris-HCl (pH 7.5) containing either 4.0×106 DWV RNA copies or 3.5 × 104 SBV RNA, between the 2nd-3rd abdominal tergites using a Harbo syringe and a disposable borosilicate needle made from modified capillary tubes. Pupae were maintained in individual wells in a 24-well plate with a piece of Whatman filter paper at the bottom of each well to prevent pupal tissue damage. In each plate, four wells were filled with sterile water for humidity, at 32°C for 7 days. Pupae were then homogenized in a 2 mL microfuge tube with a sterile, metal, 3 mm bead in 1 mL 1X PBS (pH 7.4) using the TissueLyser II (Qiagen) at 30 Hz for 2 min. Homogenates were centrifuged at 14,000 x g for 15 min at 4°C and supernatant was transferred to a fresh tube. The virus-containing lysates were filtered through a 0.45 µm filter, then a 0.22 µm filter to remove large particles and non-viral microbial contaminants, respectively. RNA was isolated from the virus-containing filtrate (20 µL) using TRIzol (Thermo) according to manufacturer’s instructions. RNA was quantified using a NanoDrop 2000c Spectrophotometer, cDNA was synthesized with reverse transcriptase, and virus RNA copies (including both genomes and transcripts) were quantified from cDNA using qPCR. The inoculum was tested for copurifying/contaminating viruses via PCR (Data S2, S2 Fig). RNA copy concentration in the purified virus stock was quantified relative to a standard curve using qPCR (Data S3). DWV genome consensus sequences were assembled from short-read sequencing data obtained from sequencing RNA isolated from the inoculum (DWV-lab 2024; GenBank PV821422) and SBV sequencing data (sacbrood virus isolate MT 2023; GenBank PV788228.1). Virus inoculum aliquots were stored at -80°C until each experiment.

Experimentally introduced honey bee virus infections

Honey bees were infected with DWV using previously described methods (72). In brief, age-matched bees were cold anesthetized at 4°C for 10 min and infected via intra-thoracic injection using a Harbo syringe with 3.5 × 104 DWV RNA copies, 3.5 × 104 SBV RNA copies, or coinfected with 3.5 × 104 DWV RNA copies and 9.1×103 SBV RNA copies. If octopamine (OA) and/or epinastine (EP) treatment was administered via injection, the injection inoculum also contained 10 mM OA and/or 4 mM EP so that each bee only experienced one injection to reduce injection-associated stress. Each inoculum dose was suspended in 2 µL10 mM Tris HCl buffer, pH 7.5. Buffer injected/ mock infected bees were injected with 2 μL buffer. The DWV doses utilized in this study were similar to previous studies and resulted in virus levels commonly found in naturally infected bees (8, 34, 37, 38, 117, 118). As previously described (33), bees had #000 steel washers affixed to their thoraces using rubber cement immediately after injections; washers had an outside diameter of 0.19 cm and weighed 4.3 mg, less than half of an average pollen load (3).

Flight mill

The flight mills used in this study were based on flight mills used in previous honey bee studies (33, 46, 119, 120) and included a magnet to tether bees from the affixed steel washer on each thorax, which enabled their removal from flight mills with minimal damage. Flight assays were performed as previously described (33). In brief, bees 72 hpi (i.e., 4-days post eclosion) were immobilized via incubation at 4°C for ∼12 minutes and tethered to flight mills; counterweights were adjusted for each bee. Temperature and humidity were maintained the same for each experiment (23°C and 31±4% relative humidity). Flight was instigated by tapping the flight arm downward and was re-initiated up to three times before bees were considered ‘exhausted’ and the data collected was representative of ‘total flight’ capability (Data S1). Bees were stored at - 80°C immediately after flight. Flight distance, duration, and speed calculations were based on redlight detector counts which were recorded on an attached Raspberry Pi as previously described (33). Active flight duration was calculated as the sum time moving >5.5 cm (one encoder count) (Data S1). Flight stop count was defined as any stop that exceeded 1.5 seconds. Honey bees may stop to seemingly engage in grooming behaviors, stops were counted if the bee re-initiated flight (33). DWV infected, EP injected, SBV infected bees fed OA or EP or injected with OA, or SBV infected bees fed both OA and EP had fewer stops than mock infected bees (Data S4).

Honey bee RNA isolation

RNA was isolated from the abdomen of each individual honey bee since abdomens were representative of the entire bee (33). Each abdomen was individually homogenized in sterile water (300 μL) with one sterile steel bead (4.5 mm) using a Qiagen TissueLyser for 2 min at 30 Hz. Lysates were centrifuged at 4°C for 5 minutes at 7,500xg. RNA was isolated from the supernatants using equal volumes of TRIzol reagent according to manufacturer’s instructions. The concentration and quality of RNA samples were assessed on a ThermoFisher Nanodrop 2000 spectrophotometer. RNA was stored at −80°C.

Reverse transcription / cDNA synthesis

Reverse transcription reactions were performed by incubating 2 μg RNA, 200 U M-MLV reverse-transcriptase and 500 ng random hexamer primers in 25 μL reactions for 1h at 37°C, according to the manufacturer’s instructions. cDNA was diluted with sterile water (1:2) and 2 μL were used for polymerase chain reactions (PCR).

Polymerase Chain Reaction

Since bees were obtained from colonies that are subject to natural infections (37, 72), PCR was performed to test for preexisting virus infections. Specifically, 2 μL cDNA template was combined with 10 pmol of each forward and reverse primers (Data S2) and amplified with ChoiceTaq polymerase according to the manufacturer’s instructions using the following conditions: 95°C for 5min; 95°C for 30s, 57°C for 30s, 72°C for 30s, 35 cycles, followed by final elongation at 72°C for 4min. PCR products were assessed by gel electrophoresis (1.5% agarose with SYBR safe) and visualized using a Syngene U:Genius 3 imaging system (S1 Fig). PCR products were previously verified by sequencing (37). Analysis of pooled cDNA samples from mock infected honey bees from each of these experiments revealed that many bees had preexisting DWV and SBV infections, therefore qPCR was used to assess their abundance in all individual samples (Data S1). The qPCR primers used in this study to target DWV, bind to both DWV-A (GenBank AY292384) and DWV-B (GenBank MN565036), in addition to numerous other DWV sequences on NCBI including the consensus sequences of the DWV inoculum as part of this study (Datal S3).

Quantitative PCR

Quantitative PCR (qPCR) was used to assess virus abundance and relative abundance of honey bee transcripts. All qPCR reactions were performed in triplicate with 2 μL of 1:2 diluted cDNA template. Each reaction contained 1× ChoiceTaq Mastermix, 0.4 μM each forward and reverse primer, 1× SYBR Green (Life Technologies), and 3 mM MgCl2 for a total volume of 20 μL per reaction. A CFX Connect Real Time instrument (BioRad) was used for the following: pre-incubation 95°C for 1 minute followed by 40 cycles of 95°C for 10s, 58°C for 20s, and 72°C for 15s, with a final melt curve analysis at 65°C for 5s to 95°C. To quantify virus copies in each sample, plasmid standards (i.e., virus-specific PCR amplicons cloned into the pGEM-T (Promega) vector and sequence verified) for each virus were used as templates, ranging from 103-109 copies per reaction for each standard curve as previously described (Data S3) (33). The honey bee gene rpl8 was used as a housekeeping gene for each sample for comparison and for relative fold change calculations (Data S1-S2). Sterile water containing no DNA template was used for negative controls. qPCR specificity was verified by melt point analysis, gel electrophoresis, and sequencing (37). The starting quantity (SQ) of cDNA template was calculated for each honey bee sample as previously described (33). For each honey bee sample, the starting quantity (SQ) of cDNA template (representing 80 ng total RNA) was calculated based on the standard curve and subtracting the average SQ of the no template control reaction (<600 copies). Samples with virus RNA copies below the limit of detection for qPCR (<1000 copies / 2 µg RNA) were listed as 0 (Data S1). Virus abundance was reported as virus RNA copies, including genomes and transcripts, per 2 µg total RNA (i.e., per RT reaction) and ranged up to 7×1010 virus RNA copies / 2µg RNA. Whole bee data was estimated based on the previous estimate of 138 µg RNA per whole bee (33). Comparison of virus abundance levels using estimated copy numbers enabled inclusion of values of 0. Using the ΔΔCt method to calculate virus levels in bees with no detectable virus would inaccurately result in ‘missing’ data and consequently skew model interpretations. Relative host gene expression was calculated using the average ΔΔCt method where ΔCt was determined by subtracting rpl8 Ct from the gene of interest Ct. Average ΔCt was calculated by the mean ΔCt of the buffer-injected bees in each experiment, which harbored the lowest virus levels. The ΔΔCt was calculated by subtracting the average control ΔCt and fold change was determined by the equation 2−ΔΔCt (Data S3).

RNA sequencing and analysis

Honey bee samples with low/no preexisting infections (n = 3-5) were selected per treatment from previously published data (33) (n=36 total) (Data S5). RNA from each sample was DNase treated using Qiagen RNeasy columns. RNA samples were sent to the Roy J. Carver Biotechnology Center at University of Illinois; quality was assessed with an Agilent 3500 fragment analyzer. Libraries were prepared via Watchmaker polyA RNA-Seq prep kit according to manufacturer instructions. Libraries were sequenced on an Illumina NovaSeq X Plus 25B (paired 150 nt reads), approximately 3.2 billion total paired end reads were obtained. Sequencing quality of reads representing individual samples using FastQC, and filtering of the comprehensive dataset was done using MultiQC versions 0.12.1 (121). Illumina adaptors were trimmed and low quality reads were omitted when Q<30 with BBduk from BBTools version 39.26 (122). Reads were normalized with shrinkage estimations for dispersion using DESeq2 l (correction with q-value<0.05) (123). Sequencing libraries were aligned to the Apis mellifera genome, Amel_HAv3.1 (i.e, GCA_003254395.2_Amel_HAv3.1_genomic.fna, downloaded from NCBI 2019–02-07) (73, 74). Sequencing data from individual samples were uploaded to the NCBI short read archive (SRA) (BioProject ID: PRJNA1414634). Genes were considered differentially expressed if log2 fold change of transcript abundance had adjusted p-values <0.05. Venn diagrams were produced using the eulerr package in R (124).

For virus quantification, trimmed paired end read files were aligned to a nonredundant version of the Holobee database (i.e., https://doi.org/10.15482/USDA.ADC/1255217) (125) which considered 32 potential virus alignments, using HISAT2 version 2.2.1 (126). Default parameters were used during alignment to allow multimapping of reads. Samples with unexpected infections were omitted from analyses (n = 6/36; Data S5). Alignment files were converted to compressed alignment files and sorted using samtools version 1.21 and unmapped reads were discarded (127). The samtools built-in idxstats command was used to quantify virus abundance in sorted alignment files and generate read counts. Read counts of each trimmed FastQ file were quantified using the Seqkit stats command (128). Computational efforts were performed on the Tempest High Performance Computing System, operated and supported by University Information Technology Research Cyberinfrastructure (RRID:SCR_026229) at Montana State University.

Statistics

Statistical analyses were performed in R 4.3.3. Log-transformations and non-parametric tests were performed as necessary. Specifically, all relative gene expression data was log2 transformed and all virus abundance data were log10(1+) transformed, which enabled inclusion of 0s. All mixed models were assessed using the lmer4 package (129). Assumptions for linear models were verified by diagnostic plots and histograms using base R functions and the performance package (130, 131). All models included experiment ID as a random effect and where relevant, virus injections and flight mill IDs were also included as random effects (Data S4). Models were compared using corrected Aikike information criterion (AICc) and Bayesian information criterion (BIC). Models were assessed via maximum likelihood (ML) when random effects varied and via restricted maximum likelihood (REML) when fixed effects varied. We reported only results from the best model for each analysis, but other considered models were included in supplementary material (Data S4). For each linear mixed model with reported estimates, p-values were estimated from t-statistics via Satterthwaite’s method (132).